---
title: "glmmTMB DE (volcanos, chr21 removed)"
format: html
---

Visualises differential expression results from the glmmTMB model fitted in `scripts/glmmTMB_no_chr21.qmd` (chromosome 21 genes excluded prior to model fitting). Produces an interaction logFC scatter plot, per-gene violin and pseudobulk boxplots, eCDF plots, and ISG/top-gene summaries.

**Inputs:** `output/run1/objects/no_chr21/glmmTMB_with_interactions_corrected.qs2`, `output/run1/objects/prelabelled_integrated_rpca.qs2`
**Outputs:** PDF figures written to `output/run1/graphs/no_chr21/`

## Load libraries

```{r}
library(Seurat)
library(glmmTMB)
library(qs2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
```

## Load directories

```{r}
run_num <- "run1"

out_dir    <- file.path("output", run_num)
graphs_dir <- file.path(out_dir, "graphs", "no_chr21")
objects_dir <- file.path(out_dir, "objects")
no_chr21_dir <- file.path(objects_dir, "no_chr21")

dir.create(graphs_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(no_chr21_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data

```{r}
required_inputs <- c(
  file.path(no_chr21_dir, "glmmTMB_with_interactions_corrected.qs2"),
  file.path(objects_dir, "prelabelled_integrated_rpca.qs2")
)
missing <- required_inputs[!file.exists(required_inputs)]
if (length(missing) > 0) {
  message("Required inputs not found:\n", paste(" -", missing, collapse = "\n"),
          "\nRun scripts/glmmTMB_no_chr21.qmd and seurat_trisomy_v1/v2.qmd first.")
  knitr::knit_exit()
}
final_df <- qs_read(required_inputs[1])

seurat_obj <- qs_read(required_inputs[2])
```

## Interaction scatter plot

```{r}
plot_data <- final_df %>%
  # 1. Keep only the rows where we are looking at the Ploidy effect (A vs B)
  # and ensure Background is one of our two target groups.
  filter(contrast == "T - D", !is.na(Background)) %>%

  # 2. Select only the necessary columns.
  # Important: Interaction_FDR must be the same for both rows of a gene
  # or you'll get duplicates again.
  select(Gene, Background, estimate, Interaction_FDR) %>%

  # 3. Pivot: this should now have 1 row per Gene
  pivot_wider(names_from = Background, values_from = estimate) %>%

  # 4. Create labeling logic
  mutate(
    Significance = ifelse(Interaction_FDR < 0.05, "Significant Interaction", "Not Significant"),
    Label = ifelse(Interaction_FDR < 0.05, Gene, NA)
  )


```

```{r}
p <- ggplot(plot_data, aes(x = NLGF, y = hAPP)) +
  # Draw the non-significant genes first so they fall in the background
  geom_point(data = filter(plot_data, Significance == "Not Significant"),
             color = "grey80", alpha = 0.5, size = 1.5) +

  # Draw your 23 significant genes on top in a bold color
  geom_point(data = filter(plot_data, Significance == "Significant Interaction"),
             color = "red", alpha = 0.9, size = 2.5) +

  # Add the "Zero" crosshairs to divide it into quadrants
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "black") +

  # Add the diagonal "Line of No Interaction" (Slope = 1, Intercept = 0)
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "darkblue", linewidth = 0.8) +

  # Add the repelling text labels for the significant genes
  geom_text_repel(aes(label = Label),
                  color = "black",
                  size = 4,
                  fontface = "bold",
                  box.padding = 0.5,
                  max.overlaps = Inf) +

  # Aesthetics
  theme_minimal() +
  labs(
    title = "Interaction Landscape: Trisomy Effect by Plaque Background (chr21 removed)",
    x = "Log2 Fold Change (Trisomic vs Disomic in NLGF)",
    y = "Log2 Fold Change (Trisomic vs Disomic in hAPP)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

print(p)
ggsave(
  file.path(graphs_dir, "interaction_logFC_scatterplot.pdf"),
  plot = p,
  width = 12,
  height = 6,
  units = "in"
  )
```

## Per-gene violin plots

```{r}
#Extract signficant interaction genes from plot data
sig_genes <- plot_data %>%
  filter(Significance == "Significant Interaction") %>%
  pull(Gene)
sig_genes


#Pull raw counts for these genes
expr_matrix <- GetAssayData(seurat_obj, assay = "SCT", layer = "data")[sig_genes, ]
expr_df <- as.data.frame(t(as.matrix(expr_matrix)))
expr_df$CellID <- rownames(expr_df)
expr_df$Ploidy <- factor(seurat_obj@meta.data$microglia, levels = c("D", "T"))
expr_df$Background <- factor(seurat_obj@meta.data$app_genotype, levels = c("NLGF", "hAPP"))
expr_df$Mouse <- seurat_obj@meta.data$orig.ident
expr_df$Condition <- factor(seurat_obj@meta.data$group_id)

# Pivot to "Long" format for ggplot2
expr_long <- expr_df %>%
  pivot_longer(cols = any_of(sig_genes), names_to = "Gene", values_to = "Expression")

pdf_file <- file.path(graphs_dir, "Interaction_Genes_Raw_Violins.pdf")
pdf(pdf_file, width = 8, height = 6)

for (gene in sig_genes) {

  # Filter data for the current gene
  gene_data <- expr_long %>% filter(Gene == gene)

  # Create the plot
  p <- ggplot(gene_data, aes(x = Condition, y = Expression, fill = Background)) +
    # scale="width" ensures all violins are the same width regardless of cell count
    geom_violin(scale = "width", alpha = 0.7, color = "black") +

    # Add a point for the mean expression to help guide the eye
    stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +

    scale_fill_manual(values = c("NLGF" = "#4C72B0", "hAPP" = "#C44E52")) +
    theme_minimal(base_size = 14) +
    labs(
      title = paste("Expression of", gene),
      subtitle = "Significant Ploidy * Background Interaction (chr21 removed)",
      x = "Condition",
      y = "Raw Counts"
    ) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      axis.text.x = element_text(face = "bold"),
      legend.position = "none"
    )

  print(p)
}

dev.off()
cat("Saved violin plots to:", pdf_file, "\n")
```

## Dotplot

```{r}
p_dot <- DotPlot(seurat_obj,
                 features = sig_genes,
                 group.by = "group_id",
                 assay = "SCT") +

  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "bold.italic"),
        axis.text.y = element_text(face = "bold", size = 12),
        plot.title = element_text(face = "bold", hjust = 0.5)) +

  scale_color_gradient(low = "lightgrey", high = "#C44E52") +

  labs(title = "Significant Ploidy * Background Interaction Genes (chr21 removed)",
       x = "Gene",
       y = "Condition")

dp_file <- file.path(graphs_dir, "Interaction_Genes_DotPlot.pdf")
pdf(dp_file, width = 12, height = 5)
print(p_dot)
dev.off()

cat("DotPlot saved to:", dp_file, "\n")
```

## Pseudobulk count plots

```{r}
# Collapse the single cells down to one average value per mouse
pb_long <- expr_long %>%
  group_by(Mouse, Condition, Background, Gene) %>%
  summarise(Pseudobulk_Expression = mean(Expression), .groups = "drop")

pdf_file <- file.path(graphs_dir, "Interaction_Genes_Pseudobulk.pdf")
pdf(pdf_file, width = 8, height = 6)

for (gene in sig_genes) {

  gene_pb <- pb_long %>% filter(Gene == gene)

  p <- ggplot(gene_pb, aes(x = Condition, y = Pseudobulk_Expression, fill = Background)) +

    geom_boxplot(alpha = 0.5, outlier.shape = NA) +

    geom_jitter(width = 0.2, size = 3, shape = 21, color = "black", stroke = 1) +

    scale_fill_manual(values = c("NLGF" = "#4C72B0", "hAPP" = "#C44E52")) +

    theme_minimal(base_size = 14) +
    labs(
      title = paste(gene, "- Pseudobulk Expression"),
      subtitle = "Each dot represents the average normalized expression of one mouse",
      x = "Condition",
      y = "Mean SCT Normalized Expression"
    ) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      axis.text.x = element_text(face = "bold", angle = 45, hjust = 1),
      legend.position = "none"
    )

  print(p)
}

dev.off()
cat("Saved pseudobulked boxplots to:", pdf_file, "\n")
```

## ECDF (non-zero cells)

```{r}

ecdf_pdf_file <- file.path(graphs_dir, "Interaction_Genes_NonZero_eCDF_with_Proportions.pdf")
pdf(ecdf_pdf_file, width = 10, height = 7)

for (gene in sig_genes) {

  gene_data_all <- expr_long %>% filter(Gene == gene)

  stats_summary <- gene_data_all %>%
    group_by(Condition) %>%
    summarise(
      total_cells = n(),
      non_zero_cells = sum(Expression > 0),
      prop_expressing = (non_zero_cells / total_cells) * 100,
      .groups = "drop"
    ) %>%
    mutate(label = paste0(Condition, ": ", round(prop_expressing, 1), "%"))

  prop_text <- paste("Expressing Cells % -", paste(stats_summary$label, collapse = " | "))

  gene_data_nonzero <- gene_data_all %>% filter(Expression > 0)

  if(nrow(gene_data_nonzero) < 5) {
    message("Skipping ", gene, " - too few expressing cells.")
    next
  }

  p_ecdf <- ggplot(gene_data_nonzero, aes(x = Expression, color = Condition)) +
    stat_ecdf(geom = "step", linewidth = 1.2) +
    theme_minimal(base_size = 12) +
    labs(
      title = paste(gene, "- Non-Zero Distribution"),
      subtitle = prop_text,
      x = "SCT Normalized Expression (Non-Zero Only)",
      y = "Cumulative Fraction of Expressing Cells",
      color = "Condition"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 9, color = "blue4"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )

  print(p_ecdf)
}

dev.off()
cat("Saved annotated eCDF plots to:", ecdf_pdf_file, "\n")
```

## ECDF (all cells)

```{r}
ecdf_pdf_file <- file.path(graphs_dir, "Interaction_Genes_AllCells_eCDF_with_Proportions.pdf")
pdf(ecdf_pdf_file, width = 10, height = 7)

for (gene in sig_genes) {

  gene_data_all <- expr_long %>% filter(Gene == gene)

  stats_summary <- gene_data_all %>%
    group_by(Condition) %>%
    summarise(
      total_cells = n(),
      non_zero_cells = sum(Expression > 0),
      prop_expressing = (non_zero_cells / total_cells) * 100,
      .groups = "drop"
    ) %>%
    mutate(label = paste0(Condition, ": ", round(prop_expressing, 1), "%"))

  prop_text <- paste("Expressing Cells % -", paste(stats_summary$label, collapse = " | "))

  if(nrow(gene_data_all) < 5) {
    message("Skipping ", gene, " - too few total cells.")
    next
  }

  p_ecdf <- ggplot(gene_data_all, aes(x = Expression, color = Condition)) +
    stat_ecdf(geom = "step", linewidth = 1.2) +
    theme_minimal(base_size = 12) +
    labs(
      title = paste(gene, "- All Cells Distribution"),
      subtitle = prop_text,
      x = "SCT Normalized Expression (All Cells)",
      y = "Cumulative Fraction of Total Cells",
      color = "Condition"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 9, color = "blue4"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )

  print(p_ecdf)
}

dev.off()
cat("Saved annotated eCDF plots to:", ecdf_pdf_file, "\n")
```

## ISG gene pseudobulk plots
```{r}
isg_genes <- c("IFI27", "IFI44L", "IFIT1", "IFIT2", "IFIT3",
               "MX1", "MX2", "OAS1", "ISG15", "USP18",
               "RSAD2", "LY6E", "DDX60", "TRIM25", "STAT1", "IRF7")

isg_nlgf <- final_df %>%
  filter(Background == "NLGF" & contrast == "T - D") %>%
  arrange(FDR) %>%
  filter(Gene %in% isg_genes) %>%
  slice_head(n = 10)

isg_happ <- final_df %>%
  filter(Background == "hAPP" & contrast == "T - D") %>%
  arrange(FDR) %>%
  filter(Gene %in% isg_genes) %>%
  slice_head(n = 10)

expr_matrix <- GetAssayData(seurat_obj, assay = "SCT", layer = "data")[isg_genes, ]
expr_df <- as.data.frame(t(as.matrix(expr_matrix)))
expr_df$CellID <- rownames(expr_df)
expr_df$Ploidy <- factor(seurat_obj@meta.data$microglia, levels = c("D", "T"))
expr_df$Background <- factor(seurat_obj@meta.data$app_genotype, levels = c("NLGF", "hAPP"))
expr_df$Mouse <- seurat_obj@meta.data$orig.ident
expr_df$Condition <- factor(seurat_obj@meta.data$group_id)

expr_long <- expr_df %>%
  pivot_longer(cols = any_of(isg_genes), names_to = "Gene", values_to = "Expression")


# Collapse the single cells down to one average value per mouse
pb_long <- expr_long %>%
  group_by(Mouse, Condition, Background, Gene) %>%
  summarise(Pseudobulk_Expression = mean(Expression), .groups = "drop")
pdf_file <- file.path(graphs_dir, "ISG_genes_Pseudobulk_glmmtmb.pdf")
pdf(pdf_file, width = 8, height = 6)

for (gene in isg_genes) {

  gene_pb <- pb_long %>% filter(Gene == gene)

  nlgf_fdr_raw <- final_df %>% filter(Gene == gene & Background == "NLGF" & contrast == "T - D") %>% pull(FDR)
  happ_fdr_raw <- final_df %>% filter(Gene == gene & Background == "hAPP" & contrast == "T - D") %>% pull(FDR)

  fmt_nlgf <- ifelse(length(nlgf_fdr_raw) > 0, signif(nlgf_fdr_raw, digits = 3), "N/A")
  fmt_happ <- ifelse(length(happ_fdr_raw) > 0, signif(happ_fdr_raw, digits = 3), "N/A")


  p <- ggplot(gene_pb, aes(x = Condition, y = Pseudobulk_Expression, fill = Background)) +

    geom_boxplot(alpha = 0.5, outlier.shape = NA) +

    geom_jitter(width = 0.2, size = 3, shape = 21, color = "black", stroke = 1) +

    scale_fill_manual(values = c("NLGF" = "#4C72B0", "hAPP" = "#C44E52")) +

    theme_minimal(base_size = 14) +
    labs(
      title = paste(gene, "- Pseudobulk Expression"),
      subtitle = paste0("FDR (NLGF): ", fmt_nlgf, "  |  FDR (hAPP): ", fmt_happ, "\nEach dot = average expression of one mouse"),
      x = "Condition",
      y = "Mean SCT Normalized Expression"
    ) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      axis.text.x = element_text(face = "bold", angle = 45, hjust = 1),
      legend.position = "none"
    )

  print(p)
}

dev.off()
cat("Saved pseudobulked boxplots to:", pdf_file, "\n")
```

## Top/bottom DE genes — NLGF trisomy effect
```{r}
pb_long <- expr_long %>%
  group_by(Mouse, Condition, Background, Gene) %>%
  summarise(Pseudobulk_Expression = mean(Expression), .groups = "drop")


top10_sig <- final_df %>%
  filter(Background == "NLGF" & contrast == "T - D") %>%
  arrange(FDR) %>%
  filter(estimate > 0) %>%
  slice_head(n = 10)

bot10_sig <- final_df %>%
  filter(Background == "NLGF" & contrast == "T - D") %>%
  arrange(FDR) %>%
  filter(estimate < 0) %>%
  slice_head(n = 10)

top_bottom_res <- bind_rows(top10_sig, bot10_sig)
target_genes <- top_bottom_res$Gene

pdf_file <- file.path(graphs_dir, "ISG_genes_top_bottom_nlgf_a_vs_b.pdf")
pdf(pdf_file, width = 8, height = 6)
for (gene in target_genes) {

  gene_pb <- pb_long %>% filter(Gene == gene)

  gene_fdr <- top_bottom_res %>% filter(Gene == gene) %>% pull(FDR)
  formatted_fdr <- signif(gene_fdr, digits = 3)

  p <- ggplot(gene_pb, aes(x = Condition, y = Pseudobulk_Expression, fill = Background)) +

    geom_boxplot(alpha = 0.5, outlier.shape = NA) +

    geom_jitter(width = 0.2, size = 3, shape = 21, color = "black", stroke = 1) +

    scale_fill_manual(values = c("NLGF" = "#4C72B0", "hAPP" = "#C44E52")) +

    theme_minimal(base_size = 14) +
    labs(
      title = paste(gene, "- Pseudobulk Expression"),
      subtitle = paste0("glmmTMB FDR: ", formatted_fdr, " | Each dot = average expression of one mouse"),
      x = "Condition",
      y = "Mean SCT Normalized Expression"
    ) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      axis.text.x = element_text(face = "bold", angle = 45, hjust = 1),
      legend.position = "none"
    )

  print(p)
}

dev.off()
cat("Saved pseudobulked boxplots to:", pdf_file, "\n")
```

## Top/bottom DE genes — hAPP trisomy effect

```{r}
pb_long <- expr_long %>%
  group_by(Mouse, Condition, Background, Gene) %>%
  summarise(Pseudobulk_Expression = mean(Expression), .groups = "drop")


top10_sig <- final_df %>%
  filter(Background == "hAPP" & contrast == "T - D") %>%
  arrange(FDR) %>%
  filter(estimate > 0) %>%
  slice_head(n = 10)

bot10_sig <- final_df %>%
  filter(Background == "hAPP" & contrast == "T - D") %>%
  arrange(FDR) %>%
  filter(estimate < 0) %>%
  slice_head(n = 10)

top_bottom_res <- bind_rows(top10_sig, bot10_sig)
target_genes <- top_bottom_res$Gene

pdf_file <- file.path(graphs_dir, "top10_bottom_happ_a_vs_b.pdf")
pdf(pdf_file, width = 8, height = 6)
for (gene in target_genes) {

  gene_pb <- pb_long %>% filter(Gene == gene)

  gene_fdr <- top_bottom_res %>% filter(Gene == gene) %>% pull(FDR)
  formatted_fdr <- signif(gene_fdr, digits = 3)

  p <- ggplot(gene_pb, aes(x = Condition, y = Pseudobulk_Expression, fill = Background)) +

    geom_boxplot(alpha = 0.5, outlier.shape = NA) +

    geom_jitter(width = 0.2, size = 3, shape = 21, color = "black", stroke = 1) +

    scale_fill_manual(values = c("NLGF" = "#4C72B0", "hAPP" = "#C44E52")) +

    theme_minimal(base_size = 14) +
    labs(
      title = paste(gene, "- Pseudobulk Expression"),
      subtitle = paste0("glmmTMB FDR: ", formatted_fdr, " | Each dot = average expression of one mouse"),
      x = "Condition",
      y = "Mean SCT Normalized Expression"
    ) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      axis.text.x = element_text(face = "bold", angle = 45, hjust = 1),
      legend.position = "none"
    )

  print(p)
}

dev.off()
cat("Saved pseudobulked boxplots to:", pdf_file, "\n")
```
