---
title: "glmmTMB Volcano Plots"
format: html
---

Volcano plots for the trisomy effect (Trisomic A vs Disomic B) in each plaque background (NLGF, hAPP), derived from the glmmTMB model fitted in `scripts/glmmTMB.qmd`. Points are coloured by direction and significance; top 20 up- and down-regulated genes are labelled.

**Inputs:** `output/run1/objects/glmmTMB_with_interactions_corrected.qs2`

## Load libraries

```{r include=TRUE, message=FALSE}
library(qs2)
library(Seurat)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

## Load data

```{r}
run_num <- "run1"
objects_dir <- file.path("output", run_num, "objects")

required_input <- file.path(objects_dir, "glmmTMB_with_interactions_corrected.qs2")
if (!file.exists(required_input)) {
  message("Required input not found: ", required_input,
          "\nRun scripts/glmmTMB.qmd first.")
  knitr::knit_exit()
}
final_df <- qs_read(required_input)

seurat_input <- file.path(objects_dir, "prelabelled_integrated_rpca.qs2")
if (!file.exists(seurat_input)) {
  message("Required input not found: ", seurat_input,
          "\nRun seurat_trisomy_v1/v2.qmd first.")
  knitr::knit_exit()
}
seurat_obj <- qs_read(seurat_input)

# avg_log_expr: per-gene mean of the SCT "data" layer (sctransform log-normalised
# counts) across all cells in prelabelled_integrated_rpca.qs2.
# Used as the A-axis (average expression) in MA plots below.
avg_expr <- rowMeans(GetAssayData(seurat_obj, assay = "SCT", layer = "data"))
final_df <- left_join(
  final_df,
  data.frame(Gene = names(avg_expr), avg_log_expr = as.numeric(avg_expr)),
  by = "Gene"
)
```

## Volcano plots — Trisomy effect (A vs B)

```{r}
volcano_cols <- c(
  "Upregulated"   = "#e41a1c",
  "Downregulated" = "#377eb8",
  "Not Significant" = "grey80"
)

generate_volcano <- function(df, bg_label) {

  dat <- df %>%
    filter(contrast == "A - B", Background == bg_label) %>%
    mutate(
      Status = case_when(
        FDR < 0.05 & estimate >  0.5 ~ "Upregulated",
        FDR < 0.05 & estimate < -0.5 ~ "Downregulated",
        TRUE ~ "Not Significant"
      )
    ) %>%
    group_by(Status) %>%
    mutate(GroupRank = rank(FDR)) %>%
    ungroup() %>%
    mutate(
      Label = case_when(
        Status != "Not Significant" & GroupRank <= 20 ~ Gene,
        TRUE ~ NA_character_
      )
    )

  n_up   <- sum(dat$Status == "Upregulated")
  n_down <- sum(dat$Status == "Downregulated")

  ggplot(dat, aes(x = estimate, y = -log10(FDR), fill = Status, color = Status)) +
    geom_point(aes(size = Status, alpha = Status), shape = 21, stroke = 0.2) +
    scale_fill_manual(values = volcano_cols) +
    scale_color_manual(values = c(
      "Upregulated"     = "black",
      "Downregulated"   = "black",
      "Not Significant" = "grey90"
    )) +
    scale_size_manual(values = c(
      "Upregulated"     = 3,
      "Downregulated"   = 3,
      "Not Significant" = 1.5
    )) +
    scale_alpha_manual(values = c(
      "Upregulated"     = 1,
      "Downregulated"   = 1,
      "Not Significant" = 0.3
    )) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed",
               color = "grey30", linewidth = 0.8) +
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed",
               color = "grey30", linewidth = 0.8) +
    geom_label_repel(
      aes(label = Label),
      fill = "white", color = "black", fontface = "bold", size = 3.5,
      box.padding = 0.5, point.padding = 0.5,
      min.segment.length = 0, segment.color = "grey50", max.overlaps = 50
    ) +
    labs(
      title    = paste0("Trisomy effect (A vs B) — ", bg_label, " background"),
      subtitle = paste0(
        "Top 20 genes labelled | FDR < 0.05 | |estimate| > 0.5",
        "  |  Up: ", n_up, "  Down: ", n_down
      ),
      x = "Estimate (log FC, glmmTMB)",
      y = bquote(bold("-Log"[10] ~ "FDR"))
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title       = element_text(face = "bold", size = 18, hjust = 0),
      plot.subtitle    = element_text(size = 12, color = "grey40", margin = margin(b = 10)),
      axis.title       = element_text(face = "bold", size = 14),
      axis.text        = element_text(face = "bold", color = "black"),
      axis.line        = element_line(color = "black", linewidth = 1),
      axis.ticks       = element_line(color = "black", linewidth = 1),
      legend.position  = "top",
      legend.title     = element_blank(),
      legend.text      = element_text(size = 12, face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey95")
    )
}
```

### NLGF background

```{r fig.width=10, fig.height=7}
generate_volcano(final_df, "NLGF")
```

### hAPP background

```{r fig.width=10, fig.height=7}
generate_volcano(final_df, "hAPP")
```

## MA plots — Trisomy effect (A vs B)

The x-axis (`avg_log_expr`) is the per-gene mean of the SCT log-normalised expression layer (`GetAssayData(assay = "SCT", layer = "data")`) across all cells in `prelabelled_integrated_rpca.qs2`. Top 20 significant genes per direction are labelled, ranked by average expression (highest first).

```{r}
generate_ma_plot <- function(df, bg_label) {

  dat <- df %>%
    filter(contrast == "A - B", Background == bg_label) %>%
    mutate(
      Status = case_when(
        FDR < 0.05 & estimate >  0.5 ~ "Upregulated",
        FDR < 0.05 & estimate < -0.5 ~ "Downregulated",
        TRUE ~ "Not Significant"
      )
    ) %>%
    group_by(Status) %>%
    mutate(GroupRank = rank(-avg_log_expr)) %>%
    ungroup() %>%
    mutate(
      Label = case_when(
        Status != "Not Significant" & GroupRank <= 20 ~ Gene,
        TRUE ~ NA_character_
      )
    )

  ggplot(dat, aes(x = avg_log_expr, y = estimate, fill = Status, color = Status)) +
    geom_point(aes(size = Status, alpha = Status), shape = 21, stroke = 0.2) +
    scale_fill_manual(values = c(
      "Upregulated"     = "#e41a1c",
      "Downregulated"   = "#377eb8",
      "Not Significant" = "grey80"
    )) +
    scale_color_manual(values = c(
      "Upregulated"     = "black",
      "Downregulated"   = "black",
      "Not Significant" = "grey90"
    )) +
    scale_size_manual(values = c(
      "Upregulated"     = 3,
      "Downregulated"   = 3,
      "Not Significant" = 1.5
    )) +
    scale_alpha_manual(values = c(
      "Upregulated"     = 1,
      "Downregulated"   = 1,
      "Not Significant" = 0.3
    )) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
    geom_hline(yintercept = c(-0.5, 0.5), linetype = "dashed", color = "grey30") +
    geom_label_repel(
      aes(label = Label),
      fill = "white", color = "black", fontface = "bold", size = 3.5,
      min.segment.length = 0, max.overlaps = 50
    ) +
    labs(
      title    = paste0("MA Plot: Trisomy effect (A vs B) — ", bg_label, " background"),
      subtitle = "Top 20 genes labelled per direction | FDR < 0.05 | |estimate| > 0.5\nx-axis: mean SCT log-normalised expression across all cells",
      x = "Mean SCT log-normalised expression (avg_log_expr)",
      y = "Estimate (log FC, glmmTMB)"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      plot.title       = element_text(face = "bold", size = 18, hjust = 0),
      plot.subtitle    = element_text(size = 12, color = "grey40", margin = margin(b = 10)),
      axis.title       = element_text(face = "bold", size = 14),
      axis.text        = element_text(face = "bold", color = "black"),
      axis.line        = element_line(color = "black", linewidth = 1),
      axis.ticks       = element_line(color = "black", linewidth = 1),
      legend.position  = "top",
      legend.title     = element_blank(),
      legend.text      = element_text(size = 12, face = "bold"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey95")
    )
}
```

### NLGF background

```{r fig.width=10, fig.height=7}
generate_ma_plot(final_df, "NLGF")
```

### hAPP background

```{r fig.width=10, fig.height=7}
generate_ma_plot(final_df, "hAPP")
```

## Top significant genes

```{r}
final_df %>%
  filter(contrast == "A - B", !is.na(Background), FDR < 0.05) %>%
  arrange(Background, FDR) %>%
  group_by(Background) %>%
  slice_head(n = 20) %>%
  select(Background, Gene, estimate, FDR, Interaction_FDR) %>%
  ungroup()
```
