---
title: "GSEA (glmmTMB, trisomy, chr21 removed)"
format: html
---

Runs gene set enrichment analysis (fgsea) on glmmTMB differential expression results from the chromosome-21-excluded model (`scripts/glmmTMB_no_chr21.qmd`). Covers four contrasts: trisomy effect in hAPP and NLGF backgrounds, and background effect in trisomic and disomic cells. Gene sets include Mancuso 2024 signatures, custom module lists, and GO/KEGG pathways.

**Inputs:** `output/run1/objects/no_chr21/glmmTMB_with_interactions_corrected.qs2`; Mancuso geneset TXTs (`../../Mancuso2024_genesets/`); Magda modules Excel; GO/KEGG `.gmt`
**Outputs:** CSV tables and barplot PDFs written to `output/run1/glmmTMB/GSEA/no_chr21/`

## Load libraries

```{r, message=FALSE}
library(SingleCellExperiment)
library(edgeR)
library(limma)
library(scuttle)
library(Seurat)
library(qs2)
library(ggrepel)
library(ggplot2)
library(stringr)
library(glue)
library(dplyr)
library(msigdbr)
library(fgsea)
library(viridis)
library(readxl)
```

## Load directories

```{r}
run_num <- "run1"

out_dir     <- file.path("output", run_num, "glmmTMB", "GSEA", "no_chr21")
graphs_dir  <- file.path(out_dir, "graphs")
objects_dir <- file.path(out_dir, "objects")

dir.create(graphs_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in data

```{r}
required_input <- file.path("output", run_num, "objects", "no_chr21", "glmmTMB_with_interactions_corrected.qs2")
if (!file.exists(required_input)) {
  message("Required input not found: ", required_input, "\nRun scripts/glmmTMB_no_chr21.qmd first.")
  knitr::knit_exit()
}
final_df <- qs_read(required_input)
```

## Prepare rank vectors

```{r}
create_glmm_rank <- function(df, target_background = NA, target_ploidy = NA, target_contrast) {

  if (!is.na(target_background)) {
    sub_df <- df %>% filter(Background == target_background & contrast == target_contrast)
  } else if (!is.na(target_ploidy)) {
    sub_df <- df %>% filter(Ploidy == target_ploidy & contrast == target_contrast)
  }

  if(nrow(sub_df) == 0) {
    stop(glue("No rows found. Check your target_background, target_ploidy, and target_contrast spelling!"))
  }

  sub_df <- sub_df %>%
    mutate(rank_metric = log2FC)

  ranks <- sub_df$rank_metric
  names(ranks) <- sub_df$Gene

  ranks <- ranks[!is.na(ranks) & !is.na(names(ranks))]
  return(sort(ranks, decreasing = TRUE))
}

comparisons_master <- list(
  "hAPP Trisomy vs Disomy"  = create_glmm_rank(final_df, target_background = "hAPP", target_contrast = "T - D"),
  "NLGF Trisomy vs Disomy"  = create_glmm_rank(final_df, target_background = "NLGF", target_contrast = "T - D"),
  "T NLGF vs T hAPP"        = create_glmm_rank(final_df, target_ploidy = "T",        target_contrast = "NLGF - hAPP"),
  "D NLGF vs D hAPP"        = create_glmm_rank(final_df, target_ploidy = "D",        target_contrast = "NLGF - hAPP")
)
```

## GSEA function

```{r}
run_gsea_analysis <- function(ranks,
                              comparison_name,
                              geneset_name,
                              pathways,
                              base_objects_dir,
                              base_graphs_dir,
                              group_A_label = "Group A",
                              group_B_label = "Group B",
                              color_A = "#EE7733",
                              color_B = "#33BBEE") {

  message(glue("Running GSEA | Set: {geneset_name} | Comp: {comparison_name}"))

  current_objects_dir <- file.path(base_objects_dir, "GSEA_Analysis", comparison_name)
  current_graphs_dir  <- file.path(base_graphs_dir, "GSEA_Analysis", comparison_name)

  dir.create(current_objects_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(current_graphs_dir, recursive = TRUE, showWarnings = FALSE)

  set.seed(42)
  fgsea_res <- suppressMessages(fgsea(
    pathways = pathways,
    stats    = ranks,
    minSize  = 15,
    maxSize  = 3000,
    nproc    = 1
  ))

  if (nrow(fgsea_res) == 0) {
    warning(glue("No significant pathways for {geneset_name} in {comparison_name}"))
    return(NULL)
  }

  fgsea_res_tidy <- fgsea_res %>%
    as_tibble() %>%
    mutate(leadingEdge = sapply(leadingEdge, paste, collapse = ",")) %>%
    mutate(pathway = gsub("_geneset", "", pathway)) %>%
    arrange(padj)

  csv_name <- glue("Table_{geneset_name}.csv")
  write.csv(fgsea_res_tidy, file.path(current_objects_dir, csv_name), row.names = FALSE)

  top_gsea <- fgsea_res_tidy %>%
    arrange(desc(NES)) %>%
    slice_head(n = 10) %>%
    bind_rows(fgsea_res_tidy %>% arrange(NES) %>% slice_head(n = 10)) %>%
    distinct(pathway, .keep_all = TRUE) %>%
    mutate(
      sig_label = case_when(
        padj < 0.001 ~ "***",
        padj < 0.01  ~ "**",
        padj < 0.05  ~ "*",
        TRUE         ~ ""
      ),
      fill_group = ifelse(NES > 0, group_A_label, group_B_label),
      pathway = factor(pathway, levels = unique(pathway[order(NES)]))
    )

  if (nrow(top_gsea) == 0) return(fgsea_res_tidy)

  axis_zoom <- 3.5

  plot_title <- glue("{sub('_.*', '', comparison_name)} ({geneset_name})")

  p <- ggplot(top_gsea, aes(x = NES, y = pathway, fill = fill_group)) +

    geom_col(width = 0.7) +

    geom_text(aes(label = sig_label, hjust = ifelse(NES > 0, 1.2, -0.2)),
              color = "white", size = 5, vjust = 0.75, fontface = "bold") +

    scale_fill_manual(values = setNames(c(color_A, color_B), c(group_A_label, group_B_label))) +

    scale_x_continuous(
      breaks = seq(-axis_zoom, axis_zoom, 0.5),
      labels = function(x) ifelse(x %% 1 == 0, x, "")
    ) +

    labs(x = "NES", y = NULL, title = plot_title) +
    theme_bw(base_size = 14) +

    theme(
      panel.grid.major.x = element_line(color = "grey92", linewidth = 0.5),
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),

      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5),
      legend.position = "none",

      plot.title = element_text(hjust = 0.5, face = "bold", size = 16, margin = margin(b = 40)),

      plot.margin = margin(t = 20, r = 10, b = 10, l = 10, unit = "pt")
    ) +

    annotate("text", x = -axis_zoom/2, y = Inf,
             label = glue("^ {group_B_label}"), color = color_B, size = 5, fontface = "bold", vjust = -1.0) +
    annotate("text", x = axis_zoom/2, y = Inf,
             label = glue("^ {group_A_label}"), color = color_A, size = 5, fontface = "bold", vjust = -1.0) +

    coord_cartesian(xlim = c(-axis_zoom, axis_zoom), clip = "off")

  plot_name <- glue("Barplot_{geneset_name}.pdf")

  dynamic_height <- 2 + (nrow(top_gsea) * 0.3)

  max_len <- max(nchar(as.character(top_gsea$pathway)))
  dynamic_width <- 5 + (max_len * 0.12)

  ggsave(filename = file.path(current_graphs_dir, plot_name),
         plot = p,
         width = dynamic_width,
         height = dynamic_height,
         limitsize = FALSE)

  print(p)

  return(fgsea_res_tidy)
}
```

## Config and gene sets

```{r}
GLOBAL_CONFIG <- list(
  output_dir = file.path(out_dir, "GSEA_Results_Master"),
  group_A    = "Group A",
  group_B    = "Group B",
  col_A      = "#EE7733",
  col_B      = "#33BBEE"
)

# 1. Mancuso (TXT files)
mancuso_dir <- "../../Mancuso2024_genesets"
mancuso_files <- list.files(path = mancuso_dir, pattern = "\\.txt$")
mancuso_list <- lapply(mancuso_files, function(x) {
  scan(file.path(mancuso_dir, x), what = "character", quiet = TRUE)
})
names(mancuso_list) <- gsub(".txt$", "", mancuso_files)

# 2. Magda (Excel)
module_data <- read_excel("/camp/home/zanettc/home/shared/zanettc/external_data/magda_modules.xlsx")
magda_list <- lapply(module_data, function(x) unique(na.omit(as.character(x))))
magda_list <- magda_list[sapply(magda_list, length) > 0]

#3. GO / KEGG
go_pathways <- gmtPathways("../../GO_human_all_genesets.gmt")
kegg_df <- msigdbr(species = "Homo sapiens", collection = "C2", subcollection = "CP:KEGG_MEDICUS")
kegg_list <- split(x = kegg_df$gene_symbol, f = kegg_df$gs_name)
go_kegg_list <- c(go_pathways, kegg_list)


gene_sets_master <- list(
  "Mancuso" = mancuso_list,
  "Magda"   = magda_list,
  "GO_KEGG" = go_kegg_list
)
```

## Execution loop

```{r}
purrr::walk2(names(gene_sets_master), gene_sets_master, function(gset_name, gset_pathways) {
  purrr::walk2(comparisons_master, names(comparisons_master), function(ranks, comp_name) {

    if (str_detect(comp_name, "Trisomy vs Disomy")) {
      label_A <- "Trisomy"
      label_B <- "Disomy"
    } else if (str_detect(comp_name, "NLGF vs.*hAPP")) {
      label_A <- "NLGF"
      label_B <- "hAPP"
    } else {
      label_A <- "Group A"
      label_B <- "Group B"
    }

    run_gsea_analysis(
      ranks            = ranks,
      comparison_name  = comp_name,
      geneset_name     = gset_name,
      pathways         = gset_pathways,
      base_objects_dir = objects_dir,
      base_graphs_dir  = graphs_dir,
      group_A_label    = label_A,
      group_B_label    = label_B,
      color_A          = GLOBAL_CONFIG$col_A,
      color_B          = GLOBAL_CONFIG$col_B
    )

    gc()
  })
})
```

## Single pathway enrichment plot

```{r}
term_to_plot <- "KEGG_MEDICUS_REFERENCE_TRANSLATION_INITIATION"

all_pathways_flat <- do.call(c, unname(gene_sets_master))

if (!term_to_plot %in% names(all_pathways_flat)) {
  stop(glue("Could not find pathway '{term_to_plot}' in gene_sets_master."))
}

pathway_genes <- all_pathways_flat[[term_to_plot]]

specific_graphs_dir <- file.path(graphs_dir, "Specific_Pathway_Plots")
dir.create(specific_graphs_dir, showWarnings = FALSE, recursive = TRUE)

purrr::walk2(comparisons_master, names(comparisons_master), function(ranks, comp_name) {

  message(glue("Plotting {term_to_plot} for {comp_name}..."))

  gsea_plot <- plotEnrichment(pathway_genes, ranks) +
    labs(
      title = term_to_plot,
      subtitle = glue("Comparison: {comp_name}"),
      x = "Rank",
      y = "Enrichment Score"
    ) +
    theme_bw(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
      panel.grid = element_blank()
    )

  ggsave(
    filename = file.path(specific_graphs_dir, glue("{term_to_plot}_{comp_name}.pdf")),
    plot = gsea_plot,
    width = 8,
    height = 5
  )
})
```

## Combined GSEA plots

### Define function

```{r}
plot_combined_gsea <- function(pathway_name, rank_list, pathways_obj, graphs_dir) {

  safe_pathway_name <- gsub("[[:punct:]]", "_", pathway_name)

  if (!pathway_name %in% names(pathways_obj)) {
    warning(glue("Pathway '{pathway_name}' not found in the provided pathway object."))
    return(NULL)
  }

  gene_set <- pathways_obj[[pathway_name]]
  plot_data_list <- list()

  for (comp_name in names(rank_list)) {

    ranks <- rank_list[[comp_name]]

    g <- suppressMessages(plotEnrichment(gene_set, ranks))

    df <- g$data
    df$Comparison <- comp_name

    plot_data_list[[comp_name]] <- df
  }

  combined_df <- bind_rows(plot_data_list)

  p <- ggplot(combined_df, aes(x = rank, y = ES, color = Comparison)) +
    geom_line(linewidth = 1.2, alpha = 0.8) +

    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +

    labs(
      title = pathway_name,
      x = "Gene Rank",
      y = "Enrichment Score"
    ) +

    scale_color_brewer(palette = "Set1") +

    theme_bw(base_size = 14) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
      panel.grid.minor = element_blank()
    )

  combined_out_dir <- file.path(graphs_dir, "Combined_GSEA_Plots")
  dir.create(combined_out_dir, showWarnings = FALSE, recursive = TRUE)

  file_name <- file.path(combined_out_dir, glue("{safe_pathway_name}_Combined.pdf"))

  ggsave(filename = file_name, plot = p, width = 8, height = 6)

  return(p)
}
```

### Combined enrichment — specific term

```{r}
term_to_plot <- "KEGG_MEDICUS_REFERENCE_TRANSLATION_INITIATION"

all_pathways_flat <- do.call(c, unname(gene_sets_master))

combined_plot <- plot_combined_gsea(
  pathway_name = term_to_plot,
  rank_list    = comparisons_master,
  pathways_obj = all_pathways_flat,
  graphs_dir   = graphs_dir
)

print(combined_plot)
```
