---
title: "hdWGCNA + glmmTMB stats (trisomy)"
format: html
---

Author: Carlo Zanetti

Tests whether WGCNA co-expression modules (from `hdwgcna_trisomy.qmd`) differ between experimental conditions using glmmTMB mixed models with a dispersion formula to account for heteroscedasticity. Produces boxplot and violin summaries with inset statistics tables.

**Inputs:** `output/run1/wgcna/objects/seurat_wgcna_basic.qs2`
**Outputs:** Stats CSVs and PDF figures written to `output/run1/wgcna/`

## Workflow overview

1\. Data Loading: Imports the Seurat object containing WGCNA results.

2\. Module Extraction: Retrieves Harmonized Module Eigengenes (hMEs) and Annotation: Renames color-based modules to biological names (e.g., "Metabolic").

3\. Pseudobulking: Aggregates single-cell ME scores into per-sample averages.

4\. Visualization & Stats: Generates boxplots with statistical testing to compare module activity across groups.

```{r}
rm(list = ls())
```

## Libraries

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(tidyverse)
library(hdWGCNA)
library(glmmTMB)
library(emmeans)
library(car)
library(qs2)
library(foreach)
library(doSNOW)
library(parallel)
library(glue)
library(cowplot)
library(ggpubr)

```

## Load directories

```{r}
run_num <- "run1"
out_dir <- file.path("output", run_num, "wgcna")
graphs_dir <- file.path(out_dir, "graphs")
objects_dir <- file.path(out_dir, "objects")
wgcna_name <- glue("emily_jan_{run_num}")

dir.create(graphs_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1. Data loading

```{r}
required_input <- file.path(objects_dir, "seurat_wgcna_basic.qs2")
if (!file.exists(required_input)) {
  message("Required input not found: ", required_input, "\nRun hdwgcna_trisomy.qmd first.")
  knitr::knit_exit()
}
seurat_obj <- qs_read(required_input)

seurat_obj$group_id <- as.factor(seurat_obj@meta.data$group_id)
```

## Correct metadata for wrongly assigned input mouse

```{r}
#BDAG56.1F was wrongly input into the metadata - should be BDAE56.1F - hu/hu not NLGF
#update name, group_id, app_genotype

table(seurat_obj@meta.data$orig.ident, seurat_obj@meta.data$group_id)

seurat_obj@meta.data <- seurat_obj@meta.data %>%
  mutate(
    app_genotype = case_when(orig.ident == "BDAG56.1F" ~ "hAPP",
                             TRUE ~ app_genotype),
    orig.ident = case_when(orig.ident == "BDAG56.1F" ~ "BDAE56.1F",
                           TRUE ~ orig.ident)
  )
seurat_obj@meta.data$group_id <- paste(seurat_obj@meta.data$app_genotype, seurat_obj@meta.data$microglia, sep = "_")
table(seurat_obj@meta.data$orig.ident, seurat_obj@meta.data$group_id)

```

## 2. Module extraction and annotation

```{r}
# 1. Get the MEs (This usually contains just the module numeric data)
hMEs <- GetMEs(seurat_obj, harmonized=TRUE)

# 2. Define your map
module_name_mapping <- c(
  "blue"      = "Blue: Metabolic",
  "turquoise" = "Turquoise: Homeostatic 2",
  "brown"     = "Brown: CRM",
  "yellow"    = "Yellow: DAM/HLA",
  "green"     = "Green: Proliferating",
  "black" = "Black: IRM",
  "red" = "Red: Homeostatic 1"
  
)

hMEs <- hMEs[, names(hMEs) %in% names(module_name_mapping)]

names(hMEs) <- module_name_mapping[names(hMEs)]

print(head(hMEs))
```
## 3. Long-form data preparation
```{r}
meta <- seurat_obj@meta.data

ME_long <- hMEs %>%
  rownames_to_column("Cell_ID") %>%
  pivot_longer(cols = -Cell_ID, names_to = "Module", values_to = "ME_Value") %>%
  left_join(
    meta %>% 
      rownames_to_column("Cell_ID") %>% 
      select(Cell_ID, Ploidy = microglia, Background = app_genotype, Mouse = orig.ident, Batch = pool_id),
    by = "Cell_ID"
  ) %>%
  mutate(
    Ploidy = factor(Ploidy),
    Background = factor(Background),
    Mouse = factor(Mouse),
    Batch = factor(Batch)
  )
```

## 4. Run glmmTMB models

```{r}
# Set contrasts for Type III ANOVA
options(contrasts = c("contr.sum", "contr.poly"))

# Initialize an empty list to store results
all_results_list <- list()

# Get unique modules
unique_modules <- unique(ME_long$Module)

cat("Running glmmTMB models for", length(unique_modules), "modules...\n")

for (mod in unique_modules) {
  cat("Processing:", mod, "\n")
  
  # Filter data for the specific module
  df_mod <- ME_long %>% filter(Module == mod)
  
  # 1. Fit the Model (Gaussian with dispersion formula for heteroscedasticity)
  model_fit <- tryCatch({
    glmmTMB(
      formula = ME_Value ~ Ploidy * Background + (1 | Mouse),
      dispformula = ~ Ploidy * Background, 
      data = df_mod, 
      family = gaussian(link="identity")
    ) 
  }, error = function(e) {
    message("Model failed for ", mod, ": ", e$message)
    return(NULL)
  }) 
  
  if (is.null(model_fit)) next
  
  # 2. Extract Omnibus Effects (Type III Anova)
  anova_res <- tryCatch({
    as.data.frame(car::Anova(model_fit, type = "III"))
  }, error = function(e) return(NULL))
  
  # 3. Extract Pairwise / Simple Effects
  emm_results <- tryCatch({
    spec1 <- pairs(emmeans(model_fit, ~ Ploidy | Background), reverse = TRUE)
    spec2 <- pairs(emmeans(model_fit, ~ Background | Ploidy), reverse = TRUE)
    rbind(as.data.frame(spec1), as.data.frame(spec2))
  }, error = function(e) return(NULL))
  
  if (is.null(emm_results) | is.null(anova_res)) next
  
  # 4. Combine Statistics
  emm_results$Module <- mod
  emm_results$Interaction_P <- anova_res["Ploidy:Background", "Pr(>Chisq)"]
  emm_results$Ploidy_Main_P <- anova_res["Ploidy", "Pr(>Chisq)"]
  emm_results$Background_Main_P <- anova_res["Background", "Pr(>Chisq)"]
  
  # Append to the list
  all_results_list[[mod]] <- emm_results
}

# Reset contrasts
options(contrasts = c("contr.treatment", "contr.poly"))

cat("All models completed.\n")
```

## FDR adjustment and final table
```{r}
# Combine list into a dataframe
final_df <- do.call(rbind, Filter(Negate(is.null), all_results_list))

# 1. FDR on the simple pairwise comparisons (A - B)
final_df <- final_df %>%
  mutate(FDR_Pairwise = p.adjust(p.value, method = "fdr"))

# 2. FDR on the Main Effects and Interaction (The "Omnibus" stats)
# We group by Module because these P-values are identical for all contrasts within a module
# 1. Create the summary with explicit column names
stats_summary <- final_df %>%
  group_by(Module) %>%
  summarise(
    Ploidy_P          = dplyr::first(Ploidy_Main_P),
    Background_P      = dplyr::first(Background_Main_P),
    Interaction_P     = dplyr::first(Interaction_P),
    .groups = "drop"
  ) %>%
  mutate(
    Ploidy_adjP      = p.adjust(Ploidy_P, method = "bonferroni"),
    Background_adjP  = p.adjust(Background_P, method = "bonferroni"),
    Interaction_adjP= p.adjust(Interaction_P, method = "bonferroni")
  )

# Join them back together
final_stats_df <- left_join(final_df, stats_summary %>% select(Module, ends_with("FDR")), by = "Module")

write.csv(final_stats_df, file.path(objects_dir, "glmmTMB_Module_Stats_Full_.csv"), row.names = FALSE)
```



## Boxplot with statistics table

```{r}
# 1. Prepare Mouse-level data (for dots/boxes)
ME_mouse <- ME_long %>%
  group_by(sample_id = Mouse, group_id = paste0(Background, "_", Ploidy), Module) %>%
  summarise(ME_Value = mean(ME_Value), .groups = "drop")

# 2. Format the Stats Table Data
table_data <- stats_summary %>%
  dplyr::select(Module, dplyr::contains("adjP")) %>% 
  dplyr::mutate(
    Module = stringr::str_replace(Module, ": ", ":\n"),
    dplyr::across(where(is.numeric), ~signif(., 2))
  )

colnames(table_data) <- gsub("adjP", "p[adj]", colnames(table_data))

# 3. Create the Table Object
tight_theme <- ttheme(
  base_size = 8, 
  padding = unit(c(4, 3), "mm"),
  colnames.style = colnames_style(parse = TRUE)
)

table_p <- ggtexttable(table_data, rows = NULL, theme = tight_theme)

# 4. Bold Significant FDR Values in the Table
for(i in 1:nrow(table_data)) {
  for(j in 2:ncol(table_data)) { 
    val <- as.numeric(table_data[i, j])
    if(!is.na(val) && val < 0.05) {
      table_p <- table_p %>% table_cell_font(row = i + 1, column = j, face = "bold")
    }
  }
}

# 5. Build the Main Plot
p_main <- ggplot(ME_mouse, aes(x = group_id, y = ME_Value, fill = group_id)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, shape = 21, size = 2.5, color = "black", stroke = 0.5) +
  facet_wrap(~Module, scales = "free_y", ncol = 4) +
  theme_cowplot() +
  labs(
    title = "Module Eigengene (glmmTMB Mixed Model Stats)",
    subtitle = "Dots represent per-mouse means; Stats account for cell-level variance & heteroscedasticity",
    y = "Module Eigengene (Mean)", 
    x = "Condition"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 10)
  )

# 6. Combine Plot and Table using cowplot
# x, y, width, and height might need slight tweaks depending on your screen resolution
final_plot <- ggdraw(p_main) + 
  draw_plot(table_p, x = 0.74, y = 0.05, width = 0.2, height = 0.45)

# 7. Save and Print
pdf(file.path(graphs_dir, "Module_Boxplots_glmmTMB_with_Inset.pdf"), width = 15, height = 10)
print(final_plot)
dev.off()

print(final_plot)
```

## Violin plot with statistics table

```{r}
# 1. Prepare Cell-level data for violins (NOT pseudobulked)
plot_data <- ME_long %>%
  mutate(group_id = factor(paste0(Background, "_", Ploidy)))

# 3. Create the Table Object with a matching "light" theme
table_theme <- ttheme("light", base_size = 9, padding = unit(c(4, 4), "mm"))
table_p <- ggtexttable(table_data, rows = NULL, theme = table_theme)

# 4. Bold Significant FDR Values in the Table
for(i in 1:nrow(table_data)) {
  for(j in 2:ncol(table_data)) {
    val <- as.numeric(table_data[i, j])
    if(!is.na(val) && val < 0.05) {
      table_p <- table_p %>% table_cell_font(row = i + 1, column = j, face = "bold")
    }
  }
}

# 5. Build the Main Plot (Violin + Boxplot)
p_main <- ggplot(plot_data, aes(x = group_id, y = ME_Value, fill = group_id)) +
  # Violin plot for single-cell distributions (transparent with colored outlines)
  geom_violin(aes(color = group_id), scale = "width", alpha = 0.4, trim = FALSE) +
  # Narrow, opaque boxplot overlaid with a solid black outline
  geom_boxplot(width = 0.15, alpha = 0.9, color = "black", outlier.shape = NA) +
  facet_wrap(~Module, scales = "free_y", ncol = 4) +
  theme_cowplot() +
  labs(
    title = "Single-Cell Module Eigengene Expression",
    subtitle = "Distributions of individual cells (glmmTMB evaluates this variance)",
    y = "Module Eigengene (Cell Level)",
    x = "Condition"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    # Grey headers for the facets
    strip.background = element_rect(fill = "grey85", color = NA),
    strip.text = element_text(face = "bold", size = 10, color = "black"),
    panel.spacing = unit(1, "lines")
  )

# 6. Combine Plot and Table using cowplot
# Positions the table exactly in the empty bottom-right slot of a 4x2 grid
final_plot <- ggdraw(p_main) +
  draw_plot(table_p, x = 0.76, y = 0.05, width = 0.23, height = 0.42)

# 7. Save and Print
pdf(file.path(graphs_dir, "Module_Violin_glmmTMB_with_Inset_bonferonni.pdf"), width = 15, height = 10)
print(final_plot)
dev.off()

print(final_plot)
```

## Save outputs
```{r}
write.csv(ME_long, file.path(objects_dir, "wgcna_cell_level_data.csv"), row.names = FALSE)
write.csv(stats_summary, file.path(objects_dir, "wgcna_stats_summary_glmmTMB.csv"), row.names = FALSE)

```

