---
title: "Seurat pipeline for Trisomy data git"
format: html
---

```{r}
rm(list = ls())
```

==================================================================================================

# Seurat analysis pipeline for trisomy dataset


====================================================================================================

## Load libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(Matrix)
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(glue)
library(readxl)
library(future)
library(gridExtra)
library(SoupX)
library(scDblFinder)
library(SingleCellExperiment)
library(qs2)
library(presto)
library(ggplot2)
library(scales)
library(ggrepel)
```

# 1. Data ingestion

## Load directories - edit for each run

```{r}
base_dir <- "/nemo/homefs/home/zanettc/home/shared/zanettc/emily_transcriptomics"

#setup output structure
#change this run number as needed
run_num <- "run1"

out_dir <- file.path(base_dir, "git_rendered", "output", run_num)
graphs_dir <- file.path(out_dir, "graphs")
objects_dir <- file.path(out_dir, "objects")

dir.create(graphs_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Re-reading in to remove clusters - loop. Ignore this on first pass!!!

```{r}
seu_h_f <- qs_read(file.path(objects_dir, "seu_h_initial_subset.qs2"))

unwanted_cells <- qs_read(file.path(objects_dir, "unwanted_cells_9_6.qs2"))
seu_h_f <- subset(seu_h_f, subset = percent.mt < 10 & scDblFinder.class == 'singlet' & !(colnames(seu_h_f) %in% unwanted_cells))
ncol(seu_h_f)

qs_save(seu_h_f, file.path(objects_dir, "seu_h_subset.qs2"))
seu_h_f <- qs_read(file.path(objects_dir, "seu_h_subset.qs2"))
```

## Naive UMAP on all data

Can be used for identifying if strictly need to integrate.

```{r}
options(future.globals.maxSize = 65 * 1024^3)
seu_h_f <- SCTransform(seu_h_f, vst.flavor="v2", verbose=FALSE)

seu_h_f <- RunPCA(seu_h_f)

seu_h_f <- RunUMAP(seu_h_f, dims=1:30)
```

### Plot of unintegrated grouped UMAPs

```{r}
p <- DimPlot(seu_h_f, 
        reduction = "umap", 
        split.by = "pool_id",
  group.by = "pool_id") +
  ggtitle("Unintegrated Data: Split by Pool")
p
ggsave(file.path(graphs_dir, "QC_umaps_pools_split.pdf"), plot = p, width = 18, height = 6, units = "in")
```

# 3. Individual batch processing

Before integrating, processed each batch individually to check for quality and assign cell type scores (e.g. Mancuso).

```{r}
batches <- SplitObject(seu_h_f, split.by = "pool_id")


options(future.globals.maxSize= 8912896000000)
batches <- lapply(batches, SCTransform)



batches <- lapply(batches, \(o) {
  set.seed(42)
  o <- RunPCA(
    o, 
    npcs = 30,
    assay = "SCT",
    verbose = FALSE
  )
  set.seed(42)
  o <- RunUMAP(
    o,
    dims = 1:25,
    n.neighbors = 30L,
    n.epochs = 500,
    min.dist = 0.05
  )
  o
})
  

invisible(lapply(names(batches), function(nm){
  p <- DimPlot(batches[[nm]], reduction = "umap", group.by = "orig.ident", shuffle=TRUE) + ggtitle(nm)
  p2<- FeaturePlot(batches[[nm]], features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), cols = c("#440154", "#FDE725"))
  p3 <- DimPlot(batches[[nm]], group.by = "scDblFinder.class") +
  ggtitle(glue("Doublet Locations Batch{nm}"))
  p4 <- DimPlot(batches[[nm]], reduction = "umap", split.by = "orig.ident", shuffle=TRUE, ncol = 3) + ggtitle(nm)
  print(p)
  print(p2)
  print(p3)
  print(p4)
  
  #create batch subdirectories for each batch 
  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  dir.create(batch_graphs_dir, recursive = TRUE, showWarnings = FALSE)
  
  ggsave(
  filename = file.path(batch_graphs_dir,glue("QC_batch{nm}_umap_samples_overlayed.pdf")),
  plot = p,
  width = 7, height = 5, dpi = 300, units = "in"
)
  ggsave(
  filename = file.path(batch_graphs_dir,glue("QC_batch{nm}_umap_qc_x3.pdf")),
  plot = p2,
  width = 7, height = 5, dpi = 300, units = "in"
)
  ggsave(
    filename = file.path(batch_graphs_dir,glue("QC_batch{nm}_doublets_umap.pdf")),
    plot = p3,
    width = 7, height = 5, dpi = 300, units = "in"
  )
  ggsave(
  filename = file.path(batch_graphs_dir,glue("QC_batch{nm}_umap_samples_split.pdf")),
  plot = p4,
  width = 7, height = 5, dpi = 300, units = "in"
)
  
  
}))

```

### Save initial clustered batches

```{r}
qs_save(batches, file = file.path(objects_dir, "umap_batches.qs2"))
```

### Read in initial data if needed

```{r}
batches <- qs_read(file.path(objects_dir, "umap_batches.qs2"))
```

## Clustering

```{r}
res = 0.2
batches <- lapply(batches, \(o) {
  set.seed(42)
  o <- FindNeighbors(o, reduction = "pca", dims = 1:30)
  set.seed(42)
  o <- FindClusters(o, resolution = res)
})

invisible(lapply(names(batches), function(nm){
  
  p <- DimPlot(batches[[nm]], reduction = "umap", label = TRUE) + ggtitle(glue("Batch {nm} UMAP {res} resolution"))
  print(p)
  
  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  
  ggsave(
  filename = file.path(batch_graphs_dir, glue("clusters_batch_{nm}_{res}.pdf")),
  plot = p,
  width = 7, height = 5, dpi = 300, units = "in")

}))
```

#### Optional save for progress

```{r}
for (i in seq_along(batches)) {
  qs_save(batches[[i]], file = paste0(objects_dir, "/", "batch", i, "_clustered.qs2"))
}
```

#### Read in if needed

```{r}
files <- list.files(
  path = objects_dir, 
  pattern = "batch[0-9]+_clustered.qs2", # Matches your specific naming pattern
  full.names = TRUE
)
batches <- lapply(files, qs_read)
```

## Initial identification of non-microglial clusters

Using signature txt files supplied by Emma

```{r}
#emma text files
sig_dir <- file.path(base_dir,"signatures")

gs_myeloid <- readLines(file.path(sig_dir, "myeloid.txt"))
myeloid_seu <- intersect(gs_myeloid, rownames(batches[[1]]))
myeloid_seu


gs_proliferating <- readLines(file.path(sig_dir, "proliferating.txt"))
prolif_seu <- intersect(gs_proliferating, rownames(batches[[1]]))
gs_proliferating

gs_macrophage <- readLines(file.path(sig_dir, "macrophage.txt"))
macrophage_seu <- intersect(gs_macrophage, rownames(batches[[1]]))
macrophage_seu

gs_dendritic <- readLines(file.path(sig_dir, "dendritic.txt"))
dendritic_seu <- intersect(gs_dendritic, rownames(batches[[1]]))
dendritic_seu


#mancuso modules
mancuso <- "/nemo/lab/destrooperb/home/shared/zanettc/Mancuso2024_genesets"  

gs_crm <- readLines(file.path(mancuso, "CRM_geneset.txt"))
gs_dam <- readLines(file.path(mancuso, "DAM_geneset.txt"))
gs_hla <- readLines(file.path(mancuso, "HLA_geneset.txt"))
gs_hm <- readLines(file.path(mancuso, "HM_geneset.txt"))
gs_irm <- readLines(file.path(mancuso, "IRM_geneset.txt"))
gs_rm <- readLines(file.path(mancuso, "RM_geneset.txt"))
gs_tcrm <- readLines(file.path(mancuso, "transCRM_geneset.txt"))

```

### Module scoring

```{r}
gene_sets <- list(
  dendritic = dendritic_seu,
  proliferating = prolif_seu,
  macrophage = macrophage_seu,
  myeloid = myeloid_seu,
  CRM = gs_crm,
  DAM = gs_dam,
  HLA = gs_hla,
  HM = gs_hm,
  IRM = gs_irm,
  RM = gs_rm,
  tCRM = gs_tcrm
)

# process each batch using an index 'i'
# seq_along() so 'i' becomes 1, 2, 3... corresponding to the list order
batches <- lapply(seq_along(batches), function(i) {
  
  # Extract the object using the index
  o <- batches[[i]]
  
  # Define a clean name for this batch
  nm <- as.character(i)
  
  message(glue("Processing Batch {nm}..."))

  # Setup Directory
  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  if (!dir.exists(batch_graphs_dir)) dir.create(batch_graphs_dir, recursive = TRUE)
  
  # 2. Add Module Scores
  # AddModuleScore calculates the average expression of the gene set 
  # subtracted by aggregated expression of control feature sets.
  o <- AddModuleScore(
    object = o, 
    features = gene_sets, 
    name = names(gene_sets),
    ctrl = 5,
    seed = 42
  )
  
  # 3. Factorize Clusters for numerical ordering
  o$seurat_clusters <- factor(
    o$seurat_clusters, 
    levels = sort(as.numeric(unique(as.character(o$seurat_clusters))))
  )
  Idents(o) <- "seurat_clusters"
  
  # 4. Generate Plots
  
  # Plot 1: Cell Types
  p1 <- VlnPlot(
    o, 
    features = c("dendritic1", "proliferating2", "macrophage3", "myeloid4"), 
    group.by = "seurat_clusters", 
    pt.size = 0, 
    ncol = 2
  ) + ggtitle(glue("Batch {nm}: Cell Types"))
  
  # Plot 2: Mancuso Signatures
  p2 <- VlnPlot(
    o, 
    features = c("CRM5", "DAM6", "HLA7", "HM8", "IRM9", "RM10", "tCRM11"), 
    group.by = "seurat_clusters", 
    pt.size = 0, 
    ncol = 4
  ) + ggtitle(glue("Batch {nm}: Mancuso Scores"))
  
  # Plot 3: QC
  p3 <- VlnPlot(
    o, 
    features = c("nFeature_RNA"), 
    group.by = "seurat_clusters", 
    pt.size = 0, 
    ncol = 2
  ) + ggtitle(glue("Batch {nm}: QC"))

  # 5. Save Plots
  ggsave(
    filename = file.path(batch_graphs_dir, glue("batch_{nm}_celltypes.pdf")),
    plot = p1, width = 7, height = 5, dpi = 300
  )
  
  ggsave(
    filename = file.path(batch_graphs_dir, glue("batch_{nm}_mancuso.pdf")),
    plot = p2, width = 10, height = 5, dpi = 300
  )
  
  ggsave(
    filename = file.path(batch_graphs_dir, glue("batch_{nm}_qc.pdf")),
    plot = p3, width = 10, height = 5, dpi = 300
  )
  
  # Return the modified object to update the list
  return(o)
})
```

# 4. Normalisation and Integration

Merges batches using reciprocal PCA (RPCA) to align them.

RPCA is relatively gentle, so removes technical batch effects while hopefully preserving biological variance.

```{r}
#merge list of batches back into one object
merged_obj <- merge(batches[[1]], y = batches[-1])

Idents(merged_obj) <- "pool_id"
```

```{r}
#run SCTransform on merged object 
options(future.globals.maxSize = 100 * 1024^3)
merged_obj <- SCTransform(merged_obj, vst.flavor = "v2", verbose = FALSE)
```

```{r}
#run PCA on unintegrated object (required for RPCA)
merged_obj <- RunPCA(merged_obj, verbose = FALSE, assay = "SCT")
```

```{r}
set.seed(42)
options(future.globals.maxSize = 100 * 1024^3)
#Change this for each diff method
int_method = "rpca"
DefaultAssay(merged_obj) <- "SCT"
merged_obj <- IntegrateLayers(
  object = merged_obj, 
  method = RPCAIntegration, #RPCAIntegration, #HarmonyIntegration
  normalization.method = "SCT", 
  orig.reduction = "pca", 
  new.reduction = glue("integrated.{int_method}"),
  verbose = FALSE
)
```

# 5. Clustering and annotation

Using integrated reduction

```{r}
set.seed(42)
merged_obj <- FindNeighbors(
  merged_obj, 
  reduction = glue("integrated.{int_method}"), 
  dims = 1:30
)
set.seed(42)
int_method <- "rpca"
cluster_resolution = 0.15

#find clusters
merged_obj <- FindClusters(
  merged_obj, 
  resolution = cluster_resolution, 
  cluster.name = glue("{int_method}_clusters"), 
  graph.name = "SCT_snn"
)
```

## Run UMAP on integrated reduction

```{r}
set.seed(42)
merged_obj <- RunUMAP(
  merged_obj, 
  #reduction = glue("integrated.{int_method}"), 
  dims = 1:30, 
  reduction.name = glue("umap.{int_method}")
)

```

## Save integrated object

```{r}
qs_save(merged_obj, file= file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))

merged_obj <- qs_read(file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))
```

## Plot diagnostic UMAP graphs

-   Initial clusters

-   Sample distribution

-   Pool distribution

-   QC (Number of genes, UMIs, % MT)

-   Doulbets and outliers

-   Split by condition

```{r}

#PLOT INITIAL CLUSTERS#########################################
p1 <- DimPlot(
  merged_obj, 
  reduction = glue("umap.{int_method}"), 
  group.by = glue("{int_method}_clusters"), 
  label = TRUE
) + ggtitle(glue("Clusters {cluster_resolution} resolution"))

#CHECK SAMPLE DISTRIBUTION#####################################
p2 <- DimPlot(
  merged_obj, 
  reduction = glue("umap.{int_method}"),  
  group.by = "orig.ident", # Or whatever column holds your batch ID
  shuffle = TRUE,          # Shuffles points so one batch doesn't hide another
  alpha = 0.5              # Transparency helps see overlap
) + ggtitle("Sample Distribution Overlaid")


p2.5 <- DimPlot(
  merged_obj, 
  reduction = glue("umap.{int_method}"),  
  split.by = "orig.ident", 
  ncol = 5,
  alpha = 0.5              # Transparency helps see overlap
) + ggtitle("Sample Distribution Split")

#CHECK POOL DISTRIBUTION#######################################
p3 <- DimPlot(
  merged_obj, 
  reduction = glue("umap.{int_method}"), 
  group.by = "pool_id", # Or whatever column holds your batch ID
  shuffle = TRUE,          # Shuffles points so one batch doesn't hide another
  alpha = 0.5              # Transparency helps see overlap
) + ggtitle("Pool Distribution")

p3.5 <- DimPlot(
  merged_obj, 
  reduction = glue("umap.{int_method}"), 
  split.by = "pool_id", # Or whatever column holds your batch ID
  shuffle = TRUE,          # Shuffles points so one batch doesn't hide another
  alpha = 0.5              # Transparency helps see overlap
) + ggtitle("Pool Distribution")

#QC OF COUNTS, # GENES, % MT TRANSCRIPTS#######################
p4 <- FeaturePlot(merged_obj, 
                  reduction = glue("umap.{int_method}"), 
                  features = c("nCount_RNA", "nFeature_RNA", "percent.mt"  ), 
                  cols = c("#440154", "#FDE725"))

#DOUBLET PLOT #################################################
p5 <- DimPlot(merged_obj,
              reduction = glue("umap.{int_method}"),
              group.by = "scDblFinder.class") +
  ggtitle("Doublet Locations")

p5.5 <- DimPlot(merged_obj,
              reduction = glue("umap.{int_method}"),
              group.by = "mad3_outlier") +
  ggtitle("MAD3 Outlier Locations")

#PLOTS FOR GROUPS #############################################
p6 <- DimPlot(merged_obj, 
              reduction = glue("umap.{int_method}"), 
              group.by = "group_id", shuffle=TRUE) +
  ggtitle("Condition Locations")

p6.5 <- DimPlot(merged_obj, 
              reduction = glue("umap.{int_method}"), 
              split.by = "group_id", ncol=5) +
  ggtitle("Condition Locations")

print(p1)
print(p2)
print(p2.5)
print(p3)
print(p4)
print(p5)
print(p5.5)
print(p6)
print(p6.5)
print(p3.5)



ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_{cluster_resolution}.pdf")),
  plot = p1,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_sample_dist_overlaid.pdf")),
  plot = p2,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_sample_dist_split.pdf")),
  plot = p2.5,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_batch_dist.pdf")),
  plot = p3,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_batch_dist_split.pdf")),
  plot = p3.5,
  width = 10, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir,int_method, cluster_resolution, glue("integrated_{int_method}_qc.pdf")),
  plot = p4,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir,int_method, cluster_resolution, glue("integrated_{int_method}_doublet_dist.pdf")),
  plot = p5,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir,int_method, cluster_resolution, glue("integrated_{int_method}_outlier_dist.pdf")),
  plot = p5.5,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_condition_dist_overlaid.pdf")),
  plot = p6,
  width = 7, height = 5, dpi = 300, units = "in")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_condition_dist_split.pdf")),
  plot = p6.5,
  width = 7, height = 5, dpi = 300, units = "in")
```

## Join layers

JoinLayers is needed to collapse separate batch raw counts into a single counts matrix, so can plot all of them together as one dataset.

For dotplots, standard log-normalisation is easier to interpret. Important as need to correct for sequencing depth.

Scale data is necessary for dotplots and heatmaps - z-scored.

```{r}
DefaultAssay(merged_obj) <- "RNA"
merged_obj <- JoinLayers(merged_obj, assay= "RNA")

#SCTransform does this automatically. 
#Corrects for some cells being sequenced deeper than others
merged_obj <- NormalizeData(merged_obj)
#some genes are louder than others - drowns out signal of actual marker genes. Scales so mean expression across all cells is 0. 
merged_obj <- ScaleData(merged_obj)
```

## Mancuso sense check

Module scores for Mancuso modules

```{r}
mancuso <- "/nemo/lab/destrooperb/home/shared/zanettc/Mancuso2024_genesets"  

gs_crm <- readLines(file.path(mancuso, "CRM_geneset.txt"))
gs_dam <- readLines(file.path(mancuso, "DAM_geneset.txt"))
gs_hla <- readLines(file.path(mancuso, "HLA_geneset.txt"))
gs_hm <- readLines(file.path(mancuso, "HM_geneset.txt"))
gs_irm <- readLines(file.path(mancuso, "IRM_geneset.txt"))
gs_rm <- readLines(file.path(mancuso, "RM_geneset.txt"))
gs_tcrm <- readLines(file.path(mancuso, "transCRM_geneset.txt"))

#Load in non-microglia gene sets
sig_dir <- "signatures"  

gs_myeloid <- readLines(file.path(base_dir, sig_dir, "myeloid.txt"))
myeloid_seu <- intersect(gs_myeloid, rownames(merged_obj))
myeloid_seu
head(rownames(merged_obj))

gs_proliferating <- readLines(file.path(base_dir,sig_dir, "proliferating.txt"))
prolif_seu <- intersect(gs_proliferating, rownames(merged_obj))
gs_proliferating

gs_macrophage <- readLines(file.path(base_dir,sig_dir, "macrophage.txt"))
macrophage_seu <- intersect(gs_macrophage, rownames(merged_obj))
macrophage_seu

gs_dendritic <- readLines(file.path(base_dir,sig_dir, "dendritic.txt"))
dendritic_seu <- intersect(gs_dendritic, rownames(merged_obj))
dendritic_seu

gene_sets <- list(
  dendritic = dendritic_seu,
  proliferating = prolif_seu,
  macrophage = macrophage_seu,
  myeloid = myeloid_seu,
  CRM = gs_crm,
  DAM = gs_dam,
  HLA = gs_hla,
  HM = gs_hm,
  IRM = gs_irm,
  RM = gs_rm,
  tCRM = gs_tcrm
)

merged_obj <- AddModuleScore(merged_obj, features = gene_sets, name = names(gene_sets))


merged_obj$seurat_clusters <- factor(
  merged_obj$rpca_clusters, 
  levels = sort(as.numeric(unique(as.character(merged_obj$rpca_clusters))))
)   
merged_obj$seurat_clusters <- Idents(merged_obj)

p <- VlnPlot(
  merged_obj, 
  features = c("dendritic1", "proliferating2", "macrophage3", "myeloid4"), 
  group.by = "seurat_clusters", 
  pt.size = 0, 
  ncol = 2
)
p

p2 <- VlnPlot(
  merged_obj, 
  features = c("CRM5", "DAM6", "HLA7", "HM8", "IRM9","RM10", "tCRM11"), 
  group.by = "seurat_clusters", 
  pt.size = 0, 
  ncol = 4
)


p3 <- VlnPlot(
  merged_obj, 
  features = c("nFeature_RNA"), 
  group.by = "seurat_clusters", 
  pt.size = 0, 
  ncol = 2
)
ggsave(
  filename = file.path(graphs_dir,int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_nonmicroglia_{cluster_resolution}.pdf")),
  plot = p,
  width = 7, height = 5, dpi = 300, units = "in"
)
ggsave(
  filename = file.path(graphs_dir,int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_mancuso_{cluster_resolution}.pdf")),
  plot = p2,
  width = 10, height = 5, dpi = 300, units = "in"
)

ggsave(
  filename = file.path(graphs_dir,int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_features_{cluster_resolution}.pdf")),
  plot = p3,
  width = 10, height = 5, dpi = 300, units = "in"
)

table(merged_obj$seurat_clusters)
```

# 6. Find cluster markers

Uses Seurat function FindAllMarkers to find the key upregulated genes for each cluster.

```{r}
DefaultAssay(merged_obj) <- "RNA"
library(presto)
Idents(merged_obj) <- glue("{int_method}_clusters")
markers <- FindAllMarkers(
  merged_obj,
  test.use = "wilcox",
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)
head(markers)

write.csv(markers, file.path(objects_dir, glue("wilcox_markers_{cluster_resolution}.csv")))

#check of top 10 cluster markers for each cluster
top10_markers <- markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

print(top10_markers)
```

## Mitochondrial % violin plot

```{r}
p <- VlnPlot(merged_obj, features = "percent.mt", pt.size = 0) + 
  geom_hline(yintercept = 15, linetype = "dashed", color = "red")

ggsave(
  filename = file.path(graphs_dir,int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_mt_{cluster_resolution}.pdf")),
  plot = p,
  width = 10, height = 5, dpi = 300, units = "in"
)
```

## Label integrated object

Labelled according to Mancuso signatures and dotplot genes.

```{r}
Idents(merged_obj)
new_cluster_ids <- c(
"0" = "HM_1",
"1" = "tCRM_1",
"2" = "tCRM_2",
"3" = "DAM/HLA",
"4" = "IRM",
"5" = "HLA/myeloid",
"6" = "Proliferating",
"7" = "HM_2", 
"8" = "CRM"
)

merged_obj <- RenameIdents(merged_obj, new_cluster_ids)

p <- DimPlot(merged_obj, reduction="umap.rpca", label=TRUE, repel=TRUE)
print(p)
ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("labelled_integrated_{cluster_resolution}.png")),
  plot = p,
  device = ragg::agg_png,  # key bit
  width = 7, height = 5, dpi = 300, units = "in"
)

```

```{r}
qs_save(merged_obj, file.path(objects_dir, glue("labelled_integrated_{cluster_resolution}.qs2")))
```

## Refining labelled umap

```{r}
library(scales)
library(ggrepel) # Required for the white text halo

# 1. Update Labels (with the fix from before) ---------------------------------
cluster_counts <- table(Idents(merged_obj))
new_labels <- paste0(names(cluster_counts), "\n n = ", comma(as.numeric(cluster_counts)))
names(new_labels) <- names(cluster_counts)
merged_obj_plot <- RenameIdents(merged_obj, new_labels)

# 2. Calculate Centroids for Custom Labeling ----------------------------------
# We need to calculate where to put the text manually to apply the 'halo' effect
umap_coords <- data.frame(Embeddings(merged_obj_plot, "umap.rpca"))
# Add the cluster identity to this dataframe
umap_coords$cluster <- Idents(merged_obj_plot)

# Calculate the mean X and Y for every cluster
centroids <- aggregate(cbind(umaprpca_1, umaprpca_2) ~ cluster, data = umap_coords, mean)

# 3. Setup Axis Geometry ------------------------------------------------------
# (Same logic as before, just ensuring we use the right column names)
x_range <- range(umap_coords[,1])
y_range <- range(umap_coords[,2])

axis_origin_x <- x_range[1] - (diff(x_range) * 0.1)
axis_origin_y <- y_range[1] - (diff(y_range) * 0.1)
axis_length   <- (diff(x_range) * 0.2)

# 4. Generate the Polished Plot -----------------------------------------------
p <- DimPlot(merged_obj_plot, 
             reduction = "umap.rpca", 
             label = FALSE,      # Turn OFF default labels
             pt.size = 0.1) +    
  NoLegend() + 
  NoAxes() + 
  theme(plot.title = element_blank()) +
  
  # --- The "Nicer" Text Layer (White Halo + Bold) ---
  geom_text_repel(data = centroids,
                  aes(x = umaprpca_1, y = umaprpca_2, label = cluster),
                  fontface = "bold",      # Bold text
                  size = 4.5,             # Slightly larger
                  color = "black",        # Text color
                  bg.color = "white",     # THE HALO: White outline
                  bg.r = 0.15,            # Radius of the halo
                  force = 5,              # Repel force (adjust if labels scatter too much)
                  min.segment.length = 0) + # Draw lines if labels are far away

  # --- L-Shaped Axes (Thicker for better look) ---
  annotate("segment", x = axis_origin_x, xend = axis_origin_x + axis_length, 
           y = axis_origin_y, yend = axis_origin_y, 
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"), 
           size = 1.2, lineend = "round") + # Thicker size (1.2)
  
  annotate("segment", x = axis_origin_x, xend = axis_origin_x, 
           y = axis_origin_y, yend = axis_origin_y + axis_length, 
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"), 
           size = 1.2, lineend = "round") +

  # --- Axis Labels (Bold) ---
  annotate("text", x = axis_origin_x + (axis_length/2), y = axis_origin_y - 1, 
           label = "UMAP 1", hjust = 0.5, vjust = 1, fontface = "bold") +
  annotate("text", x = axis_origin_x - 1, y = axis_origin_y + (axis_length/2), 
           label = "UMAP 2", angle = 90, hjust = 1, vjust = 0.5, fontface = "bold") +

  # --- Total Count (Bold) ---
  annotate("text", x = mean(x_range), y = axis_origin_y - 2.5, 
           label = paste0("Total cells = ", comma(length(Cells(merged_obj)))), 
           size = 5, fontface = "bold")

print(p)

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("labelled_polished_{cluster_resolution}.pdf")),
  plot = p,
  device = cairo_pdf,
  width = 8, height = 6, dpi = 300, units = "in"
)
```

```{r}
sessionInfo()
```


