---
title: "Seurat pipeline for Trisomy data — Part 1: QC, Filtering and Initial Integration"
author: "Carlo Zanetti"
format: html
---

```{r}
rm(list = ls())
```

# Seurat analysis pipeline for trisomy dataset

This script processes 10x Genomics scRNA-seq data derived from xenografted human microglia into mouse models. It covers the full pipeline from raw CellRanger outputs through to initial cluster identification and removal of non-microglial populations. Outputs from this script are read in by Part 2.

## Workflow overview

1.  **Data ingestion**: Loads CellRanger outputs and maps to metadata
2.  **QC and filtering**:
    -   Verifies file integrity
    -   Separates human vs mouse cells based on gene prefixes
    -   Checkpoint save — first initial object
    -   Labels multiplets (scDblFinder) and outliers (MAD score)
3.  **Individual batch processing** — QC and module scoring per pool
4.  **Normalisation and Integration**:
    -   Merges sequencing pools
    -   Normalises using SCTransform v2
    -   Corrects batch effects using RPCA
5.  **Clustering and annotation**:
    -   Generates UMAPs
    -   Scores cells against curated Mancuso signatures
6.  **Cluster markers**
7.  **Identification and removal of non-microglial clusters**

## Load libraries

```{r}
#| message: false
library(dplyr)
library(Matrix)
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(glue)
library(readxl)
library(future)
library(gridExtra)
library(SoupX)
library(scDblFinder)
library(SingleCellExperiment)
library(qs2)
library(presto)
library(ggplot2)
library(scales)
library(ggrepel)
```

# 1. Data ingestion

## Load directories

Edit `base_dir` and `run_num` for each new run. Output directories are created automatically if they do not already exist.

```{r}
base_dir <- "/nemo/homefs/home/zanettc/home/shared/zanettc/emily_transcriptomics"
demux_path <- file.path(base_dir, "input/jan_raw/cellranger_output")

nums <- c("1", "2", "3", "4", "5", "6")

filtered_paths <- file.path(
    unlist(lapply(
      file.path(demux_path, glue("pool{nums}_out"), "outs", "per_sample_outs"),
      list.dirs, recursive = FALSE, full.names = TRUE
    )),
    "count", "sample_filtered_feature_bc_matrix"
  )

raw_paths <- file.path(
    unlist(lapply(
      file.path(demux_path, glue("pool{nums}_out"), "outs", "per_sample_outs"),
      list.dirs, recursive = FALSE, full.names = TRUE
    )),
    "count", "sample_raw_feature_bc_matrix"
  )

run_num <- "run1"

out_dir     <- file.path("output", run_num)
graphs_dir  <- file.path(out_dir, "graphs")
objects_dir <- file.path(out_dir, "objects")

dir.create(graphs_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in metadata

```{r}
meta <- read_excel(file.path(base_dir, "input/meta/meta_data_combined.xlsx"))
```

## Size check

Verifies that the three expected CellRanger output files exist and reports their sizes for each sample.

```{r}
expected <- c("matrix.mtx.gz", "features.tsv.gz", "barcodes.tsv.gz")

check_files_by_path <- function(path) {
  sample_id   <- basename(dirname(dirname(path)))
  lane_folder <- basename(dirname(dirname(dirname(dirname(dirname(path))))))
  pool_number <- as.integer(gsub("\\D", "", lane_folder))

  files       <- file.path(path, expected)
  file_present <- file.exists(files)
  sizes       <- ifelse(file_present, file.info(files)$size, NA_real_)

  tibble(
    pool_id     = pool_number,
    sample_hash = sample_id,
    file        = expected,
    exists      = file_present,
    size        = sizes
  )
}

results <- map_dfr(filtered_paths, check_files_by_path)
results
```

## Map sample hashes to mouse IDs

Each sample hash returned by CellRanger is matched against the experimental metadata to recover the biological mouse ID and pool assignment.

```{r}
path_map <- results %>%
  select(pool_id, sample_hash) %>%
  distinct() %>%
  mutate(
    filtered_path = filtered_paths,
    raw_path      = raw_paths
  )

named_paths_tbl <- path_map %>%
  left_join(meta, by = c("sample_hash", "pool_id"))

filtered_data_dirs <- named_paths_tbl %>%
  filter(!is.na(mouse_id)) %>%  # drops blank tag (B0255) used by Emma in cellranger
  select(mouse_id, filtered_path) %>%
  deframe()

raw_data_dirs <- named_paths_tbl %>%
  filter(!is.na(mouse_id)) %>%
  select(mouse_id, raw_path) %>%
  deframe()
```

# 2. QC and filtering

## Removal of failed samples

These samples were removed due to poor CellRanger QC metrics or high reported cell death from FACS.

```{r}
bad_samples <- c("BDAE56.1B", "BDAG185.1G", "BDAE50.1D")

filter_samples <- function(data_dirs) {
  data_dirs[!names(data_dirs) %in% bad_samples]
}
filtered_data_dirs <- filter_samples(filtered_data_dirs)
raw_data_dirs      <- filter_samples(raw_data_dirs)
```

## Read 10x data

Loading CellRanger output with `Read10X` requires all antibody capture panels to match exactly across samples. Because last year's samples used a different hash antibody (Hash5), a direct merge crashes. Instead, each sample is loaded individually — extracting only the Gene Expression modality — before merging into a single Seurat object. Cell barcodes are prefixed with their mouse ID at merge to ensure uniqueness. A global gene filter retains only genes detected in at least 3 cells.

```{r}
make_seurat_object <- function(path, name) {
  data      <- Read10X(data.dir = path)
  rna_counts <- data$`Gene Expression`

  CreateSeuratObject(
    counts      = rna_counts,
    project     = name,
    min.features = 200
  )
}

seurat_list <- mapply(
  FUN      = make_seurat_object,
  path     = filtered_data_dirs,
  name     = names(filtered_data_dirs),
  SIMPLIFY = FALSE
)

seurat_obj <- merge(
  x           = seurat_list[[1]],
  y           = seurat_list[-1],
  add.cell.ids = names(filtered_data_dirs)
)

seurat_obj <- JoinLayers(seurat_obj)

seurat_obj@meta.data <- seurat_obj@meta.data %>%
  rownames_to_column("cell") %>%
  left_join(meta, by = c("orig.ident" = "mouse_id")) %>%
  column_to_rownames("cell")

seurat_obj@meta.data$group_id <- paste(
  seurat_obj@meta.data$app_genotype,
  seurat_obj@meta.data$microglia,
  sep = "_"
)

counts     <- GetAssayData(seurat_obj, assay = "RNA", layer = "counts")
keep_genes <- Matrix::rowSums(counts > 0) >= 3
seu        <- subset(seurat_obj, features = names(which(keep_genes)))
```

## Filter for human cells and genes

Cells are assigned a species identity based on the fraction of UMIs mapping to human (GRCh38) vs mouse (GRCm39) genes. Cells with ≥90% human UMIs are retained. Mouse genes are then dropped from the feature set, and the GRCh38 prefix is stripped from remaining gene IDs for downstream compatibility.

```{r}
ids      <- rownames(seu)
is_human <- grepl("^GRCh38", ids)
is_mouse <- grepl("^GRCm39", ids)

human_ids <- ids[is_human]
mouse_ids <- ids[is_mouse]

raw_mat      <- GetAssayData(seu, layer = "counts")
human_counts <- Matrix::colSums(raw_mat[human_ids, , drop = FALSE])
mouse_counts <- Matrix::colSums(raw_mat[mouse_ids, , drop = FALSE])
total_counts <- Matrix::colSums(raw_mat)

seu$human_frac <- ifelse(total_counts > 0, human_counts / total_counts, 0)
seu$mouse_frac <- ifelse(total_counts > 0, mouse_counts / total_counts, 0)

human_cut <- 0.90
mouse_cut <- 0.90
seu$species_call <- dplyr::case_when(
  seu$human_frac >= human_cut ~ "human",
  seu$mouse_frac >= mouse_cut ~ "mouse",
  TRUE                        ~ "mixed_or_ambiguous"
)
print(table(seu$species_call))

filter_species <- function(seurat_obj, species = "human") {
  subset(seurat_obj, subset = species_call == species)
}
seu_h <- filter_species(seu, "human")

keep_features <- rownames(seu_h)[!grepl("^GRCm39-", rownames(seu_h))]
seu_h         <- subset(seu_h, features = keep_features)

rownames(seu_h) <- sub("^GRCh38-", "", rownames(seu_h))

dup <- any(duplicated(rownames(seu_h))); if (dup) warning("Duplicated gene IDs after renaming")
```

## Initial save

```{r}
qs_save(seu_h, file.path(objects_dir, "seu_h_initial.qs2"))
seu_h <- qs_read(file.path(objects_dir, "seu_h_initial.qs2"))
```

## Doublet detection using scDblFinder

Multiplets are identified using [scDblFinder](https://www.10xgenomics.com/support/universal-three-prime-gene-expression/documentation/steps/library-prep/chromium-gem-x-single-cell-3-v4-gene-expression-user-guide). The expected doublet rate is set to 0.4% per 1,000 recovered cells, consistent with the v4 10x chemistry specification. Detection is run per sequencing pool to account for pool-level variation.

```{r}
sce <- as.SingleCellExperiment(seu_h)

# dbr.per1k: for every 1,000 recovered cells, add 0.4% to the expected doublet rate (v4 chemistry)
sce <- scDblFinder(sce, samples = "pool_id", dbr.per1k = 0.004)

seu_h$scDblFinder.class    <- sce$scDblFinder.class
seu_h$scDblFinder.score    <- sce$scDblFinder.score
seu_h$scDblFinder.weighted <- sce$scDblFinder.weighted

table(seu_h$pool_id, seu_h$scDblFinder.class)

qs_save(seu_h, file.path(objects_dir, "seu_h_scdbl.qs2"))
```

### Doublet score histogram

```{r}
p <- ggplot(seu_h@meta.data, aes(x = scDblFinder.score, fill = scDblFinder.class)) +
  geom_histogram(bins = 100, position = "identity", alpha = 0.6) +
  scale_x_log10() +
  theme_classic() +
  labs(
    title    = "Doublet Score Distribution",
    subtitle = "Look for a valley between Singlets (Grey) \nand Doublets (Red)",
    x        = "scDblFinder Score (Log Scale)",
    y        = "Count"
  ) +
  scale_fill_manual(values = c("singlet" = "grey", "doublet" = "red"))

p
ggsave(file.path(graphs_dir, "QC_scdbl_hist.pdf"), plot = p, width = 12, height = 6, units = "in")
```

### QC violin plot

Violin plots of UMI counts, feature counts, and mitochondrial percentage split by pool, coloured by scDblFinder classification.

```{r}
seu_h[["percent.mt"]] <- PercentageFeatureSet(seu_h, pattern = "^MT-")

p <- VlnPlot(
  seu_h,
  features  = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
  group.by  = "pool_id",
  split.by  = "scDblFinder.class",
  pt.size   = 0
) +
  labs(fill = "Cell Class") +
  theme(legend.position = "right")
p
ggsave(file.path(graphs_dir, "QC_scDbl_violin.pdf"), plot = p, width = 12, height = 6, units = "in")
```

### Counts vs features

```{r}
plot1 <- FeatureScatter(seu_h, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(seu_h, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

ggsave(file.path(graphs_dir, "QC_nCount_vs_mt.pdf"),      plot = plot1, width = 12, height = 6, units = "in")
ggsave(file.path(graphs_dir, "QC_nCount_vs_feature.pdf"), plot = plot2, width = 12, height = 6, units = "in")

print(plot1)
print(plot2)
```

## Counts vs features coloured by doublet assignment

```{r}
plot <- FeatureScatter(
  seu_h,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA",
  group.by = "scDblFinder.class"
)
ggsave(file.path(graphs_dir, "QC_nCount_vs_feature_doublets.pdf"), plot = plot, width = 12, height = 6, units = "in")
print(plot)
```

## Outlier detection (MAD scoring)

Outliers are flagged using the Median Absolute Deviation (MAD) rather than fixed thresholds, making the cutoff robust to pool-level differences in sequencing depth. Per-pool statistics are calculated and cutoffs at both 3 and 5 MADs above the median are retained for comparison.

```{r}
seu_h@meta.data <- seu_h@meta.data %>%
  select(-any_of(c("median_count", "mad_count", "upper_limit_5_mad", "upper_limit_3_mad")))

qc_stats <- seu_h@meta.data %>%
  group_by(pool_id) %>%
  summarise(
    median_count = median(nCount_RNA),
    mad_count    = mad(nCount_RNA)
  ) %>%
  mutate(
    upper_limit_5_mad = median_count + (5 * mad_count),
    upper_limit_3_mad = median_count + (3 * mad_count)
  )

meta_updated <- seu_h@meta.data %>%
  rownames_to_column("cell_id") %>%
  left_join(qc_stats, by = "pool_id") %>%
  mutate(
    mad5_outlier = ifelse(nCount_RNA >= upper_limit_5_mad, "Outlier", "Not Outlier"),
    mad3_outlier = ifelse(nCount_RNA >= upper_limit_3_mad, "Outlier", "Not Outlier")
  ) %>%
  column_to_rownames("cell_id")

seu_h@meta.data <- meta_updated

VlnPlot(seu_h, features = "nCount_RNA", group.by = "pool_id", pt.size = 0.1) +
  geom_hline(data = qc_stats, aes(yintercept = upper_limit_5_mad), color = "red", linetype = "dashed")
```

### Intersection of MAD outliers and doublet calls

Calculates the number and percentage of cells that would be removed at each MAD threshold, combined with doublet removal.

```{r}
dbl_col <- "scDblFinder.class"

global_stats <- seu_h@meta.data %>%
  summarise(
    total_cells      = n(),
    n_remove_3       = sum(mad3_outlier == "Outlier" | .data[[dbl_col]] == "doublet"),
    percent_remove_3 = round((n_remove_3 / total_cells) * 100, 2),
    n_remove_5       = sum(mad5_outlier == "Outlier" | .data[[dbl_col]] == "doublet"),
    percent_remove_5 = round((n_remove_5 / total_cells) * 100, 2)
  )

print(global_stats)
```

### Outlier vs doublet co-occurrence

```{r}
table(Outlier_Status = seu_h$mad3_outlier, Doublet_Status = seu_h$scDblFinder.class)
```

### Cells with high feature counts within outlier groups

Checks how many cells exceed 6,000 features within each MAD category, as a sanity check against previously used fixed cutoffs.

```{r}
outlier_summary <- seu_h@meta.data %>%
  group_by(mad3_outlier, mad5_outlier) %>%
  summarise(
    Total_Cells              = n(),
    Cells_Over_6000_Features = sum(nFeature_RNA > 6000),
    Percentage_High_Feature  = round((Cells_Over_6000_Features / Total_Cells) * 100, 2),
    .groups = "drop"
  )

outlier_summary
```

### MAD cutoff comparison

Prints the number of cells retained or lost at each threshold and overlays both cutoff lines on a violin plot.

```{r}
counts <- seu_h$nCount_RNA

median_val <- median(counts)
mad_val    <- mad(counts)

cutoff_3mad <- median_val + (3 * mad_val)
cutoff_5mad <- median_val + (5 * mad_val)

cat("Median UMI Count:", median_val, "\n")
cat("MAD:", mad_val, "\n\n")

cat("--- 3 MAD Cutoff ---\n")
cat("Threshold:", round(cutoff_3mad, 2), "\n")
cat("Cells removed:", sum(counts > cutoff_3mad), "\n")
cat("Cells kept:",   sum(counts <= cutoff_3mad), "\n\n")

cat("--- 5 MAD Cutoff ---\n")
cat("Threshold:", round(cutoff_5mad, 2), "\n")
cat("Cells removed:", sum(counts > cutoff_5mad), "\n")
cat("Cells kept:",   sum(counts <= cutoff_5mad), "\n")

p <- VlnPlot(seu_h, features = "nCount_RNA", pt.size = 0, group.by = "pool_id") +
  geom_hline(yintercept = cutoff_3mad, color = "red",  linetype = "dashed", size = 1) +
  geom_hline(yintercept = cutoff_5mad, color = "blue", linetype = "dashed", size = 1) +
  annotate("text", x = 5.1, y = 100000,
           label = paste0("3 MAD: ", round(cutoff_3mad)),
           vjust = 1.5, hjust = 0, color = "red", fontface = "bold") +
  annotate("text", x = 5.1, y = 100000,
           label = paste0("5 MAD: ", round(cutoff_5mad)),
           vjust = -0.5, hjust = 0, color = "blue", fontface = "bold") +
  ggtitle("nCount_RNA Distribution: 3 vs 5 MAD") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Pool number") +
  ylab("UMI counts")

ggsave(
  filename = file.path(graphs_dir, glue("count_dist_3_vs_5_mad.pdf")),
  plot = p, width = 8, height = 5, dpi = 300, units = "in"
)
```

### Count vs feature scatter with MAD cutoffs

```{r}
qc_scatter <- FeatureScatter(seu_h, feature1 = "nCount_RNA", feature2 = "nFeature_RNA",
                             group.by = "pool_id", shuffle = TRUE) +
  geom_vline(xintercept = cutoff_3mad, color = "red",  linetype = "dashed", size = 1) +
  annotate("text", x = cutoff_3mad, y = max(seu_h$nFeature_RNA), label = "3 MAD",
           color = "red",  angle = 90, vjust = -0.5, hjust = 1, fontface = "bold") +
  geom_vline(xintercept = cutoff_5mad, color = "blue", linetype = "dashed", size = 1) +
  annotate("text", x = cutoff_5mad, y = max(seu_h$nFeature_RNA), label = "5 MAD",
           color = "blue", angle = 90, vjust = -0.5, hjust = 1, fontface = "bold") +
  ggtitle("Count vs Feature (with MAD Cutoffs)") +
  xlab("Number of counts") +
  ylab("Number of genes")

ggsave(
  filename = file.path(graphs_dir, "QC_count_vs_feat_lines.pdf"),
  plot = qc_scatter, width = 7, height = 5, dpi = 300, units = "in"
)
print(qc_scatter)
```

## Initial subsetting

Cells are filtered on three criteria: fewer than 500 detected genes, mitochondrial content above 15%, or flagged as a MAD5 outlier. Doublets are retained here and removed in Part 2 once cluster identity has been confirmed.

```{r}
ncol(seu_h)
seu_h_f <- subset(seu_h, subset = nFeature_RNA > 500 & percent.mt < 15 & mad5_outlier != "Outlier")
ncol(seu_h_f)
qs_save(seu_h_f, file.path(objects_dir, "seu_h_initial_subset.qs2"))
```

## Naive UMAP on all data

A quick unintegrated UMAP gives an initial view of structure and reveals whether pool-level batch effects require correction.

```{r}
options(future.globals.maxSize = 65 * 1024^3)
seu_h_f <- SCTransform(seu_h_f, vst.flavor = "v2", verbose = FALSE)
seu_h_f <- RunPCA(seu_h_f)
seu_h_f <- RunUMAP(seu_h_f, dims = 1:30)
```

### Unintegrated UMAP split by pool

```{r}
p <- DimPlot(seu_h_f,
             reduction = "umap",
             split.by  = "pool_id",
             group.by  = "pool_id") +
  ggtitle("Unintegrated Data: Split by Pool")
p
ggsave(file.path(graphs_dir, "QC_umaps_pools_split.pdf"), plot = p, width = 18, height = 6, units = "in")
```

# 3. Individual batch processing

Each pool is processed independently to assess quality and score cell type identity before integration. This allows per-pool QC inspection and ensures Mancuso signatures are applied at the batch level.

## Load gene signatures

Gene sets for non-microglial cell types (provided by Emma) and Mancuso 2024 microglial states are loaded here for use in both batch-level and integrated scoring.

```{r}
sig_dir <- file.path(base_dir, "signatures")

gs_myeloid       <- readLines(file.path(sig_dir, "myeloid.txt"))
gs_proliferating <- readLines(file.path(sig_dir, "proliferating.txt"))
gs_macrophage    <- readLines(file.path(sig_dir, "macrophage.txt"))
gs_dendritic     <- readLines(file.path(sig_dir, "dendritic.txt"))

mancuso_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/Mancuso2024_genesets"

gs_crm  <- readLines(file.path(mancuso_dir, "CRM_geneset.txt"))
gs_dam  <- readLines(file.path(mancuso_dir, "DAM_geneset.txt"))
gs_hla  <- readLines(file.path(mancuso_dir, "HLA_geneset.txt"))
gs_hm   <- readLines(file.path(mancuso_dir, "HM_geneset.txt"))
gs_irm  <- readLines(file.path(mancuso_dir, "IRM_geneset.txt"))
gs_rm   <- readLines(file.path(mancuso_dir, "RM_geneset.txt"))
gs_tcrm <- readLines(file.path(mancuso_dir, "transCRM_geneset.txt"))
```

## Run per-batch UMAP and clustering

Each batch is normalised with SCTransform, reduced with PCA, and embedded with UMAP. Clustering at resolution 0.2 generates initial cluster assignments for module scoring.

```{r}
batches <- SplitObject(seu_h_f, split.by = "pool_id")

options(future.globals.maxSize = 8912896000000)
batches <- lapply(batches, SCTransform)

batches <- lapply(batches, \(o) {
  set.seed(42)
  o <- RunPCA(o, npcs = 30, assay = "SCT", verbose = FALSE)
  set.seed(42)
  o <- RunUMAP(o, dims = 1:25, n.neighbors = 30L, n.epochs = 500, min.dist = 0.05)
  o
})

invisible(lapply(names(batches), function(nm) {
  p  <- DimPlot(batches[[nm]], reduction = "umap", group.by = "orig.ident", shuffle = TRUE) + ggtitle(nm)
  p2 <- FeaturePlot(batches[[nm]], features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), cols = c("#440154", "#FDE725"))
  p3 <- DimPlot(batches[[nm]], group.by = "scDblFinder.class") + ggtitle(glue("Doublet Locations Batch {nm}"))
  p4 <- DimPlot(batches[[nm]], reduction = "umap", split.by = "orig.ident", shuffle = TRUE, ncol = 3) + ggtitle(nm)

  print(p); print(p2); print(p3); print(p4)

  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  dir.create(batch_graphs_dir, recursive = TRUE, showWarnings = FALSE)

  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_umap_samples_overlayed.pdf")), plot = p,  width = 7, height = 5, dpi = 300, units = "in")
  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_umap_qc_x3.pdf")),             plot = p2, width = 7, height = 5, dpi = 300, units = "in")
  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_doublets_umap.pdf")),           plot = p3, width = 7, height = 5, dpi = 300, units = "in")
  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_umap_samples_split.pdf")),      plot = p4, width = 7, height = 5, dpi = 300, units = "in")
}))
```

### Save / reload batch objects

```{r}
qs_save(batches, file = file.path(objects_dir, "umap_batches.qs2"))
```

```{r}
batches <- qs_read(file.path(objects_dir, "umap_batches.qs2"))
```

## Clustering

```{r}
res <- 0.2
batches <- lapply(batches, \(o) {
  set.seed(42)
  o <- FindNeighbors(o, reduction = "pca", dims = 1:30)
  set.seed(42)
  o <- FindClusters(o, resolution = res)
})

invisible(lapply(names(batches), function(nm) {
  p <- DimPlot(batches[[nm]], reduction = "umap", label = TRUE) +
    ggtitle(glue("Batch {nm} UMAP {res} resolution"))
  print(p)

  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  ggsave(
    filename = file.path(batch_graphs_dir, glue("clusters_batch_{nm}_{res}.pdf")),
    plot = p, width = 7, height = 5, dpi = 300, units = "in"
  )
}))
```

#### Optional save / reload for clustered batches

```{r}
for (i in seq_along(batches)) {
  qs_save(batches[[i]], file = paste0(objects_dir, "/batch", i, "_clustered.qs2"))
}
```

```{r}
files   <- list.files(path = objects_dir, pattern = "batch[0-9]+_clustered.qs2", full.names = TRUE)
batches <- lapply(files, qs_read)
```

## Module scoring per batch

Gene sets are intersected against the batch feature space before scoring. `AddModuleScore` calculates average expression of each gene set relative to a randomly sampled control set, giving a background-corrected enrichment score per cell.

```{r}
myeloid_seu    <- intersect(gs_myeloid,       rownames(batches[[1]]))
prolif_seu     <- intersect(gs_proliferating, rownames(batches[[1]]))
macrophage_seu <- intersect(gs_macrophage,    rownames(batches[[1]]))
dendritic_seu  <- intersect(gs_dendritic,     rownames(batches[[1]]))

gene_sets <- list(
  dendritic     = dendritic_seu,
  proliferating = prolif_seu,
  macrophage    = macrophage_seu,
  myeloid       = myeloid_seu,
  CRM           = gs_crm,
  DAM           = gs_dam,
  HLA           = gs_hla,
  HM            = gs_hm,
  IRM           = gs_irm,
  RM            = gs_rm,
  tCRM          = gs_tcrm
)

batches <- lapply(seq_along(batches), function(i) {
  o  <- batches[[i]]
  nm <- as.character(i)
  message(glue("Processing Batch {nm}..."))

  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  if (!dir.exists(batch_graphs_dir)) dir.create(batch_graphs_dir, recursive = TRUE)

  o <- AddModuleScore(object = o, features = gene_sets, name = names(gene_sets), ctrl = 5, seed = 42)

  o$seurat_clusters <- factor(
    o$seurat_clusters,
    levels = sort(as.numeric(unique(as.character(o$seurat_clusters))))
  )
  Idents(o) <- "seurat_clusters"

  p1 <- VlnPlot(o, features = c("dendritic1", "proliferating2", "macrophage3", "myeloid4"),
                group.by = "seurat_clusters", pt.size = 0, ncol = 2) +
    ggtitle(glue("Batch {nm}: Cell Types"))
  p2 <- VlnPlot(o, features = c("CRM5", "DAM6", "HLA7", "HM8", "IRM9", "RM10", "tCRM11"),
                group.by = "seurat_clusters", pt.size = 0, ncol = 4) +
    ggtitle(glue("Batch {nm}: Mancuso Scores"))
  p3 <- VlnPlot(o, features = c("nFeature_RNA"),
                group.by = "seurat_clusters", pt.size = 0, ncol = 2) +
    ggtitle(glue("Batch {nm}: QC"))

  print(p1); print(p2); print(p3)

  ggsave(filename = file.path(batch_graphs_dir, glue("batch_{nm}_celltypes.pdf")), plot = p1, width = 7,  height = 5, dpi = 300)
  ggsave(filename = file.path(batch_graphs_dir, glue("batch_{nm}_mancuso.pdf")),   plot = p2, width = 10, height = 5, dpi = 300)
  ggsave(filename = file.path(batch_graphs_dir, glue("batch_{nm}_qc.pdf")),        plot = p3, width = 10, height = 5, dpi = 300)

  return(o)
})
```

# 4. Normalisation and Integration

Batches are merged back into a single object and re-normalised with SCTransform v2. PCA is then run on the merged, unintegrated object to provide the input embedding required by RPCA. Integration with `IntegrateLayers` (RPCA method) corrects pool-level batch effects while aiming to preserve biological variance.

```{r}
merged_obj <- merge(batches[[1]], y = batches[-1])
Idents(merged_obj) <- "pool_id"

options(future.globals.maxSize = 100 * 1024^3)
merged_obj <- SCTransform(merged_obj, vst.flavor = "v2", verbose = FALSE)
merged_obj <- RunPCA(merged_obj, verbose = FALSE, assay = "SCT")

set.seed(42)
int_method <- "rpca"
DefaultAssay(merged_obj) <- "SCT"
merged_obj <- IntegrateLayers(
  object               = merged_obj,
  method               = RPCAIntegration,
  normalization.method = "SCT",
  orig.reduction       = "pca",
  new.reduction        = glue("integrated.{int_method}"),
  verbose              = FALSE
)
```

# 5. Clustering and annotation

Neighbours are found on the integrated RPCA reduction and clusters are identified at resolution 0.2. A UMAP is computed on the same integrated reduction for visualisation.

```{r}
set.seed(42)
merged_obj <- FindNeighbors(merged_obj, reduction = glue("integrated.{int_method}"), dims = 1:30)

set.seed(42)
cluster_resolution <- 0.2
merged_obj <- FindClusters(
  merged_obj,
  resolution   = cluster_resolution,
  cluster.name = glue("{int_method}_clusters"),
  graph.name   = "SCT_snn"
)
```

## Run UMAP on integrated reduction

```{r}
set.seed(42)
merged_obj <- RunUMAP(
  merged_obj,
  dims           = 1:30,
  reduction.name = glue("umap.{int_method}")
)
```

## Save integrated object

```{r}
qs_save(merged_obj, file = file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))
merged_obj <- qs_read(file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))
```

## Diagnostic UMAP plots

A standard set of diagnostic plots checks cluster structure, sample and pool mixing, QC metric distributions, doublet and outlier locations, and condition separation.

```{r}
p1 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = glue("{int_method}_clusters"), label = TRUE) +
  ggtitle(glue("Clusters {cluster_resolution} resolution"))

p2 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "orig.ident", shuffle = TRUE, alpha = 0.5) +
  ggtitle("Sample Distribution Overlaid")

p2.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                split.by = "orig.ident", ncol = 5, alpha = 0.5) +
  ggtitle("Sample Distribution Split")

p3 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "pool_id", shuffle = TRUE, alpha = 0.5) +
  ggtitle("Pool Distribution")

p3.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                split.by = "pool_id", shuffle = TRUE, alpha = 0.5) +
  ggtitle("Pool Distribution")

p4 <- FeaturePlot(merged_obj, reduction = glue("umap.{int_method}"),
                  features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
                  cols = c("#440154", "#FDE725"))

p5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "scDblFinder.class") +
  ggtitle("Doublet Locations")

p5.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                group.by = "mad3_outlier") +
  ggtitle("MAD3 Outlier Locations")

p6 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "group_id", shuffle = TRUE) +
  ggtitle("Condition Locations")

p6.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                split.by = "group_id", ncol = 5) +
  ggtitle("Condition Locations")

print(p1); print(p2); print(p2.5)
print(p3); print(p4); print(p5)
print(p5.5); print(p6); print(p6.5); print(p3.5)

ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_{cluster_resolution}.pdf")),       plot = p1,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_sample_dist_overlaid.pdf")),                plot = p2,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_sample_dist_split.pdf")),                   plot = p2.5, width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_batch_dist.pdf")),                          plot = p3,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_batch_dist_split.pdf")),                    plot = p3.5, width = 10, height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_qc.pdf")),                                  plot = p4,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_doublet_dist.pdf")),                        plot = p5,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_outlier_dist.pdf")),                        plot = p5.5, width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_condition_dist_overlaid.pdf")),              plot = p6,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_condition_dist_split.pdf")),                 plot = p6.5, width = 7,  height = 5, dpi = 300, units = "in")
```

## Join layers and normalise

`JoinLayers` collapses the separate per-pool raw count matrices into a single layer, required for downstream functions. Log-normalisation corrects for differences in sequencing depth, and `ScaleData` z-scores expression so that highly expressed genes do not dominate dotplots and heatmaps.

```{r}
DefaultAssay(merged_obj) <- "RNA"
merged_obj <- JoinLayers(merged_obj, assay = "RNA")
merged_obj <- NormalizeData(merged_obj)
merged_obj <- ScaleData(merged_obj)
```

## Mancuso sense check

Module scores are calculated for all Mancuso microglial states and non-microglial cell types on the integrated object. Gene sets are re-intersected here against the merged object feature space. Scores are plotted per cluster to guide annotation.

```{r}
myeloid_seu    <- intersect(gs_myeloid,       rownames(merged_obj))
prolif_seu     <- intersect(gs_proliferating, rownames(merged_obj))
macrophage_seu <- intersect(gs_macrophage,    rownames(merged_obj))
dendritic_seu  <- intersect(gs_dendritic,     rownames(merged_obj))

gene_sets <- list(
  dendritic     = dendritic_seu,
  proliferating = prolif_seu,
  macrophage    = macrophage_seu,
  myeloid       = myeloid_seu,
  CRM           = gs_crm,
  DAM           = gs_dam,
  HLA           = gs_hla,
  HM            = gs_hm,
  IRM           = gs_irm,
  RM            = gs_rm,
  tCRM          = gs_tcrm
)

merged_obj <- AddModuleScore(merged_obj, features = gene_sets, name = names(gene_sets))

merged_obj$seurat_clusters <- factor(
  merged_obj$rpca_clusters,
  levels = sort(as.numeric(unique(as.character(merged_obj$rpca_clusters))))
)
merged_obj$seurat_clusters <- Idents(merged_obj)

p  <- VlnPlot(merged_obj, features = c("dendritic1", "proliferating2", "macrophage3", "myeloid4"),
              group.by = "seurat_clusters", pt.size = 0, ncol = 2)
p2 <- VlnPlot(merged_obj, features = c("CRM5", "DAM6", "HLA7", "HM8", "IRM9", "RM10", "tCRM11"),
              group.by = "seurat_clusters", pt.size = 0, ncol = 4)
p3 <- VlnPlot(merged_obj, features = c("nFeature_RNA"),
              group.by = "seurat_clusters", pt.size = 0, ncol = 2)

ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_nonmicroglia_{cluster_resolution}.pdf")), plot = p,  width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_mancuso_{cluster_resolution}.pdf")),      plot = p2, width = 10, height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_features_{cluster_resolution}.pdf")),     plot = p3, width = 10, height = 5, dpi = 300, units = "in")

print(p); print(p2); print(p3)
table(merged_obj$seurat_clusters)
```

# 6. Cluster markers

Marker genes are identified per cluster using a Wilcoxon rank-sum test. Only positive markers are returned, filtered to a minimum 25% detection rate and log2FC ≥ 0.25. The top 10 markers per cluster by log2FC are printed.

```{r}
DefaultAssay(merged_obj) <- "RNA"
Idents(merged_obj) <- glue("{int_method}_clusters")
markers <- FindAllMarkers(
  merged_obj,
  test.use      = "wilcox",
  only.pos      = TRUE,
  min.pct       = 0.25,
  logfc.threshold = 0.25
)

write.csv(markers, file.path(objects_dir, glue("wilcox_markers_{cluster_resolution}.csv")))

top10_markers <- markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

top10_markers
```

## Mitochondrial % per cluster

```{r}
p <- VlnPlot(merged_obj, features = "percent.mt", pt.size = 0) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_mt_{cluster_resolution}.pdf")),
  plot = p, width = 10, height = 5, dpi = 300, units = "in"
)
print(p)
```

# 7. Removal of unwanted clusters

Clusters 9 and 6 were identified as non-microglial based on elevated mitochondrial content and non-microglial module scores. Their cell barcodes are saved here for removal in Part 2.

```{r}
unwanted_cells <- WhichCells(merged_obj, idents = c("9", "6"))
length(unwanted_cells)
qs_save(unwanted_cells, file.path(objects_dir, "unwanted_cells_9_6.qs2"))
```

```{r}
sessionInfo()
```
