---
title: "Seurat pipeline for Trisomy data — Part 2: Final Filtering, Re-integration and Cluster Labelling"
author: "Carlo Zanetti"
format: html
---

```{r}
rm(list = ls())
```

# Seurat analysis pipeline for trisomy dataset

This script is the second pass of the pipeline and continues from Part 1. It reads the initial filtered object and the list of unwanted cluster barcodes identified in Part 1, applies final cell-level filters, then re-runs normalisation, integration, clustering, and cluster labelling on the clean dataset.

**Inputs required from Part 1:**
- `seu_h_initial_subset.qs2` — initial QC-filtered object
- `unwanted_cells_9_6.qs2` — barcodes for non-microglial clusters to remove

## Load libraries

```{r setup}
#| message: false
library(dplyr)
library(Matrix)
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(glue)
library(readxl)
library(future)
library(gridExtra)
library(SoupX)
library(scDblFinder)
library(SingleCellExperiment)
library(qs2)
library(presto)
library(ggplot2)
library(scales)
library(ggrepel)
```

# 1. Setup

## Load directories

Edit `base_dir` and `run_num` to match the paths used in Part 1.

```{r}
base_dir <- "/nemo/homefs/home/zanettc/home/shared/zanettc/emily_transcriptomics"

run_num <- "run1"

out_dir     <- file.path("output", run_num)
graphs_dir  <- file.path(out_dir, "graphs")
objects_dir <- file.path(out_dir, "objects")

dir.create(graphs_dir,  recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Final filtering and cluster removal

The initial subset object from Part 1 is loaded and three additional filters are applied: cells with mitochondrial content ≥10%, doublets identified by scDblFinder, and barcodes belonging to the non-microglial clusters identified in Part 1 (clusters 9 and 6) are all removed. The result is saved as the analysis-ready object.

```{r}
required_inputs <- c(
  file.path(objects_dir, "seu_h_initial_subset.qs2"),
  file.path(objects_dir, "unwanted_cells_9_6.qs2")
)
missing <- required_inputs[!file.exists(required_inputs)]
if (length(missing) > 0) {
  message("Required inputs not found:\n", paste(" -", missing, collapse = "\n"),
          "\nRun seurat_trisomy_v1.qmd first.")
  knitr::knit_exit()
}
seu_h_f <- qs_read(required_inputs[1])

unwanted_cells <- qs_read(required_inputs[2])
seu_h_f <- subset(
  seu_h_f,
  subset = percent.mt < 10 &
           scDblFinder.class == "singlet" &
           !(colnames(seu_h_f) %in% unwanted_cells)
)
ncol(seu_h_f)

qs_save(seu_h_f, file.path(objects_dir, "seu_h_subset.qs2"))
seu_h_f <- qs_read(file.path(objects_dir, "seu_h_subset.qs2"))
```

## Naive UMAP on filtered data

A quick unintegrated UMAP gives an initial view of structure on the clean dataset and confirms whether pool-level batch effects are still present.

```{r}
options(future.globals.maxSize = 65 * 1024^3)
seu_h_f <- SCTransform(seu_h_f, vst.flavor = "v2", verbose = FALSE)
seu_h_f <- RunPCA(seu_h_f)
seu_h_f <- RunUMAP(seu_h_f, dims = 1:30)
```

### Unintegrated UMAP split by pool

```{r}
p <- DimPlot(seu_h_f,
             reduction = "umap",
             split.by  = "pool_id",
             group.by  = "pool_id") +
  ggtitle("Unintegrated Data: Split by Pool")
p
ggsave(file.path(graphs_dir, "QC_umaps_pools_split.pdf"), plot = p, width = 18, height = 6, units = "in")
```

# 2. Individual batch processing

Each pool is processed independently to confirm quality on the clean data and re-score cell type identity before integration.

## Load gene signatures

Gene sets for non-microglial cell types (provided by Emma) and Mancuso 2024 microglial states are loaded here for use in both batch-level and integrated scoring.

```{r}
sig_dir <- file.path(base_dir, "signatures")

gs_myeloid       <- readLines(file.path(sig_dir, "myeloid.txt"))
gs_proliferating <- readLines(file.path(sig_dir, "proliferating.txt"))
gs_macrophage    <- readLines(file.path(sig_dir, "macrophage.txt"))
gs_dendritic     <- readLines(file.path(sig_dir, "dendritic.txt"))

mancuso_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/Mancuso2024_genesets"

gs_crm  <- readLines(file.path(mancuso_dir, "CRM_geneset.txt"))
gs_dam  <- readLines(file.path(mancuso_dir, "DAM_geneset.txt"))
gs_hla  <- readLines(file.path(mancuso_dir, "HLA_geneset.txt"))
gs_hm   <- readLines(file.path(mancuso_dir, "HM_geneset.txt"))
gs_irm  <- readLines(file.path(mancuso_dir, "IRM_geneset.txt"))
gs_rm   <- readLines(file.path(mancuso_dir, "RM_geneset.txt"))
gs_tcrm <- readLines(file.path(mancuso_dir, "transCRM_geneset.txt"))
```

## Run per-batch UMAP and clustering

Each batch is normalised with SCTransform, reduced with PCA, and embedded with UMAP. Clustering at resolution 0.2 generates initial cluster assignments for module scoring.

```{r}
batches <- SplitObject(seu_h_f, split.by = "pool_id")

options(future.globals.maxSize = 8912896000000)
batches <- lapply(batches, SCTransform)

batches <- lapply(batches, \(o) {
  set.seed(42)
  o <- RunPCA(o, npcs = 30, assay = "SCT", verbose = FALSE)
  set.seed(42)
  o <- RunUMAP(o, dims = 1:25, n.neighbors = 30L, n.epochs = 500, min.dist = 0.05)
  o
})

invisible(lapply(names(batches), function(nm) {
  p  <- DimPlot(batches[[nm]], reduction = "umap", group.by = "orig.ident", shuffle = TRUE) + ggtitle(nm)
  p2 <- FeaturePlot(batches[[nm]], features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), cols = c("#440154", "#FDE725"))
  p3 <- DimPlot(batches[[nm]], group.by = "scDblFinder.class") + ggtitle(glue("Doublet Locations Batch {nm}"))
  p4 <- DimPlot(batches[[nm]], reduction = "umap", split.by = "orig.ident", shuffle = TRUE, ncol = 3) + ggtitle(nm)

  print(p); print(p2); print(p3); print(p4)

  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  dir.create(batch_graphs_dir, recursive = TRUE, showWarnings = FALSE)

  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_umap_samples_overlayed.pdf")), plot = p,  width = 7, height = 5, dpi = 300, units = "in")
  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_umap_qc_x3.pdf")),             plot = p2, width = 7, height = 5, dpi = 300, units = "in")
  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_doublets_umap.pdf")),           plot = p3, width = 7, height = 5, dpi = 300, units = "in")
  ggsave(filename = file.path(batch_graphs_dir, glue("QC_batch{nm}_umap_samples_split.pdf")),      plot = p4, width = 7, height = 5, dpi = 300, units = "in")
}))
```

### Save / reload batch objects

```{r}
qs_save(batches, file = file.path(objects_dir, "umap_batches.qs2"))
```

```{r}
batches <- qs_read(file.path(objects_dir, "umap_batches.qs2"))
```

## Clustering

```{r}
res <- 0.2
batches <- lapply(batches, \(o) {
  set.seed(42)
  o <- FindNeighbors(o, reduction = "pca", dims = 1:30)
  set.seed(42)
  o <- FindClusters(o, resolution = res)
})

invisible(lapply(names(batches), function(nm) {
  p <- DimPlot(batches[[nm]], reduction = "umap", label = TRUE) +
    ggtitle(glue("Batch {nm} UMAP {res} resolution"))
  print(p)

  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  ggsave(
    filename = file.path(batch_graphs_dir, glue("clusters_batch_{nm}_{res}.pdf")),
    plot = p, width = 7, height = 5, dpi = 300, units = "in"
  )
}))
```

#### Optional save / reload for clustered batches

```{r}
for (i in seq_along(batches)) {
  qs_save(batches[[i]], file = paste0(objects_dir, "/batch", i, "_clustered.qs2"))
}
```

```{r}
files   <- list.files(path = objects_dir, pattern = "batch[0-9]+_clustered.qs2", full.names = TRUE)
batches <- lapply(files, qs_read)
```

## Module scoring per batch

Gene sets are intersected against the batch feature space before scoring. `AddModuleScore` calculates average expression of each gene set relative to a randomly sampled control set, giving a background-corrected enrichment score per cell.

```{r}
myeloid_seu    <- intersect(gs_myeloid,       rownames(batches[[1]]))
prolif_seu     <- intersect(gs_proliferating, rownames(batches[[1]]))
macrophage_seu <- intersect(gs_macrophage,    rownames(batches[[1]]))
dendritic_seu  <- intersect(gs_dendritic,     rownames(batches[[1]]))

gene_sets <- list(
  dendritic     = dendritic_seu,
  proliferating = prolif_seu,
  macrophage    = macrophage_seu,
  myeloid       = myeloid_seu,
  CRM           = gs_crm,
  DAM           = gs_dam,
  HLA           = gs_hla,
  HM            = gs_hm,
  IRM           = gs_irm,
  RM            = gs_rm,
  tCRM          = gs_tcrm
)

batches <- lapply(seq_along(batches), function(i) {
  o  <- batches[[i]]
  nm <- as.character(i)
  message(glue("Processing Batch {nm}..."))

  batch_graphs_dir <- file.path(graphs_dir, glue("batch_{nm}"))
  if (!dir.exists(batch_graphs_dir)) dir.create(batch_graphs_dir, recursive = TRUE)

  o <- AddModuleScore(object = o, features = gene_sets, name = names(gene_sets), ctrl = 5, seed = 42)

  o$seurat_clusters <- factor(
    o$seurat_clusters,
    levels = sort(as.numeric(unique(as.character(o$seurat_clusters))))
  )
  Idents(o) <- "seurat_clusters"

  p1 <- VlnPlot(o, features = c("dendritic1", "proliferating2", "macrophage3", "myeloid4"),
                group.by = "seurat_clusters", pt.size = 0, ncol = 2) +
    ggtitle(glue("Batch {nm}: Cell Types"))
  p2 <- VlnPlot(o, features = c("CRM5", "DAM6", "HLA7", "HM8", "IRM9", "RM10", "tCRM11"),
                group.by = "seurat_clusters", pt.size = 0, ncol = 4) +
    ggtitle(glue("Batch {nm}: Mancuso Scores"))
  p3 <- VlnPlot(o, features = c("nFeature_RNA"),
                group.by = "seurat_clusters", pt.size = 0, ncol = 2) +
    ggtitle(glue("Batch {nm}: QC"))

  ggsave(filename = file.path(batch_graphs_dir, glue("batch_{nm}_celltypes.pdf")), plot = p1, width = 7,  height = 5, dpi = 300)
  ggsave(filename = file.path(batch_graphs_dir, glue("batch_{nm}_mancuso.pdf")),   plot = p2, width = 10, height = 5, dpi = 300)
  ggsave(filename = file.path(batch_graphs_dir, glue("batch_{nm}_qc.pdf")),        plot = p3, width = 10, height = 5, dpi = 300)

  return(o)
})
```

# 3. Normalisation and Integration

Batches are merged back into a single object and re-normalised with SCTransform v2. PCA is run on the merged, unintegrated object to provide the input embedding required by RPCA. Integration with `IntegrateLayers` (RPCA method) corrects pool-level batch effects while aiming to preserve biological variance.

```{r}
merged_obj <- merge(batches[[1]], y = batches[-1])
Idents(merged_obj) <- "pool_id"

options(future.globals.maxSize = 100 * 1024^3)
merged_obj <- SCTransform(merged_obj, vst.flavor = "v2", verbose = FALSE)
merged_obj <- RunPCA(merged_obj, verbose = FALSE, assay = "SCT")

set.seed(42)
int_method <- "rpca"
DefaultAssay(merged_obj) <- "SCT"
merged_obj <- IntegrateLayers(
  object               = merged_obj,
  method               = RPCAIntegration,
  normalization.method = "SCT",
  orig.reduction       = "pca",
  new.reduction        = glue("integrated.{int_method}"),
  verbose              = FALSE
)
```

# 4. Clustering and annotation

Neighbours are found on the integrated RPCA reduction and clusters are identified at resolution 0.15. A UMAP is computed on the same integrated reduction for visualisation.

```{r}
set.seed(42)
merged_obj <- FindNeighbors(merged_obj, reduction = glue("integrated.{int_method}"), dims = 1:30)

set.seed(42)
cluster_resolution <- 0.15
merged_obj <- FindClusters(
  merged_obj,
  resolution   = cluster_resolution,
  cluster.name = glue("{int_method}_clusters"),
  graph.name   = "SCT_snn"
)
```

## Run UMAP on integrated reduction

```{r}
set.seed(42)
merged_obj <- RunUMAP(
  merged_obj,
  dims           = 1:30,
  reduction.name = glue("umap.{int_method}")
)
```

## Save integrated object

```{r}
qs_save(merged_obj, file = file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))
merged_obj <- qs_read(file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))
```

## Diagnostic UMAP plots

A standard set of diagnostic plots checks cluster structure, sample and pool mixing, QC metric distributions, doublet and outlier locations, and condition separation.

```{r}
p1 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = glue("{int_method}_clusters"), label = TRUE) +
  ggtitle(glue("Clusters {cluster_resolution} resolution"))

p2 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "orig.ident", shuffle = TRUE, alpha = 0.5) +
  ggtitle("Sample Distribution Overlaid")

p2.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                split.by = "orig.ident", ncol = 5, alpha = 0.5) +
  ggtitle("Sample Distribution Split")

p3 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "pool_id", shuffle = TRUE, alpha = 0.5) +
  ggtitle("Pool Distribution")

p3.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                split.by = "pool_id", shuffle = TRUE, alpha = 0.5) +
  ggtitle("Pool Distribution")

p4 <- FeaturePlot(merged_obj, reduction = glue("umap.{int_method}"),
                  features = c("nCount_RNA", "nFeature_RNA", "percent.mt"),
                  cols = c("#440154", "#FDE725"))

p5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "scDblFinder.class") +
  ggtitle("Doublet Locations")

p5.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                group.by = "mad3_outlier") +
  ggtitle("MAD3 Outlier Locations")

p6 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
              group.by = "group_id", shuffle = TRUE) +
  ggtitle("Condition Locations")

p6.5 <- DimPlot(merged_obj, reduction = glue("umap.{int_method}"),
                split.by = "group_id", ncol = 5) +
  ggtitle("Condition Locations")

print(p1); print(p2); print(p2.5)
print(p3); print(p4); print(p5)
print(p5.5); print(p6); print(p6.5); print(p3.5)

ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_{cluster_resolution}.pdf")),       plot = p1,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_sample_dist_overlaid.pdf")),                plot = p2,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_sample_dist_split.pdf")),                   plot = p2.5, width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_batch_dist.pdf")),                          plot = p3,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_batch_dist_split.pdf")),                    plot = p3.5, width = 10, height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_qc.pdf")),                                  plot = p4,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_doublet_dist.pdf")),                        plot = p5,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_outlier_dist.pdf")),                        plot = p5.5, width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_condition_dist_overlaid.pdf")),              plot = p6,   width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_condition_dist_split.pdf")),                 plot = p6.5, width = 7,  height = 5, dpi = 300, units = "in")
```

## Join layers and normalise

`JoinLayers` collapses the separate per-pool raw count matrices into a single layer, required for downstream functions. Log-normalisation corrects for differences in sequencing depth, and `ScaleData` z-scores expression so that highly expressed genes do not dominate dotplots and heatmaps.

```{r}
DefaultAssay(merged_obj) <- "RNA"
merged_obj <- JoinLayers(merged_obj, assay = "RNA")
merged_obj <- NormalizeData(merged_obj)
merged_obj <- ScaleData(merged_obj)
```

## Mancuso sense check

Module scores are calculated for all Mancuso microglial states and non-microglial cell types on the integrated object. Gene sets are re-intersected here against the merged object feature space. Scores are plotted per cluster to confirm that annotations from Part 1 hold on the clean dataset.

```{r}
myeloid_seu    <- intersect(gs_myeloid,       rownames(merged_obj))
prolif_seu     <- intersect(gs_proliferating, rownames(merged_obj))
macrophage_seu <- intersect(gs_macrophage,    rownames(merged_obj))
dendritic_seu  <- intersect(gs_dendritic,     rownames(merged_obj))

gene_sets <- list(
  dendritic     = dendritic_seu,
  proliferating = prolif_seu,
  macrophage    = macrophage_seu,
  myeloid       = myeloid_seu,
  CRM           = gs_crm,
  DAM           = gs_dam,
  HLA           = gs_hla,
  HM            = gs_hm,
  IRM           = gs_irm,
  RM            = gs_rm,
  tCRM          = gs_tcrm
)

merged_obj <- AddModuleScore(merged_obj, features = gene_sets, name = names(gene_sets))

merged_obj$seurat_clusters <- factor(
  merged_obj$rpca_clusters,
  levels = sort(as.numeric(unique(as.character(merged_obj$rpca_clusters))))
)
merged_obj$seurat_clusters <- Idents(merged_obj)

p  <- VlnPlot(merged_obj, features = c("dendritic1", "proliferating2", "macrophage3", "myeloid4"),
              group.by = "seurat_clusters", pt.size = 0, ncol = 2)
p2 <- VlnPlot(merged_obj, features = c("CRM5", "DAM6", "HLA7", "HM8", "IRM9", "RM10", "tCRM11"),
              group.by = "seurat_clusters", pt.size = 0, ncol = 4)
p3 <- VlnPlot(merged_obj, features = c("nFeature_RNA"),
              group.by = "seurat_clusters", pt.size = 0, ncol = 2)

ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_nonmicroglia_{cluster_resolution}.pdf")), plot = p,  width = 7,  height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_mancuso_{cluster_resolution}.pdf")),      plot = p2, width = 10, height = 5, dpi = 300, units = "in")
ggsave(filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_features_{cluster_resolution}.pdf")),     plot = p3, width = 10, height = 5, dpi = 300, units = "in")

print(p); print(p2); print(p3)
table(merged_obj$seurat_clusters)
```

# 5. Cluster markers

Marker genes are identified per cluster using a Wilcoxon rank-sum test. Only positive markers are returned, filtered to a minimum 25% detection rate and log2FC ≥ 0.25. The top 10 markers per cluster by log2FC are printed.

```{r}
DefaultAssay(merged_obj) <- "RNA"
Idents(merged_obj) <- glue("{int_method}_clusters")
markers <- FindAllMarkers(
  merged_obj,
  test.use        = "wilcox",
  only.pos        = TRUE,
  min.pct         = 0.25,
  logfc.threshold = 0.25
)

write.csv(markers, file.path(objects_dir, glue("wilcox_markers_{cluster_resolution}.csv")))

top10_markers <- markers %>%
  group_by(cluster) %>%
  slice_max(n = 10, order_by = avg_log2FC)

print(top10_markers)
```

## Mitochondrial % per cluster

```{r}
p <- VlnPlot(merged_obj, features = "percent.mt", pt.size = 0) +
  geom_hline(yintercept = 15, linetype = "dashed", color = "red")

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("integrated_{int_method}_clusters_vs_mt_{cluster_resolution}.pdf")),
  plot = p, width = 10, height = 5, dpi = 300, units = "in"
)
```

# 6. Cluster labelling

## Label integrated object

Cluster labels are assigned based on Mancuso microglial state signatures and top marker genes identified in the previous section. Labels are applied to the Seurat identity slot for downstream plotting.

```{r}
Idents(merged_obj)
new_cluster_ids <- c(
  "0" = "HM_1",
  "1" = "tCRM_1",
  "2" = "tCRM_2",
  "3" = "DAM/HLA",
  "4" = "IRM",
  "5" = "HLA/myeloid",
  "6" = "Proliferating",
  "7" = "HM_2",
  "8" = "CRM"
)

merged_obj <- RenameIdents(merged_obj, new_cluster_ids)

p <- DimPlot(merged_obj, reduction = "umap.rpca", label = TRUE, repel = TRUE)
print(p)
ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("labelled_integrated_{cluster_resolution}.png")),
  plot = p,
  device = ragg::agg_png,
  width = 7, height = 5, dpi = 300, units = "in"
)
```

```{r}
qs_save(merged_obj, file.path(objects_dir, glue("labelled_integrated_{cluster_resolution}.qs2")))
```

## Polished UMAP

The standard `DimPlot` output is refined with custom L-shaped axis arrows, bold halo labels positioned at cluster centroids, and a total cell count annotation. This produces a publication-ready version of the labelled embedding.

```{r}
cluster_counts <- table(Idents(merged_obj))
new_labels     <- paste0(names(cluster_counts), "\n n = ", comma(as.numeric(cluster_counts)))
names(new_labels) <- names(cluster_counts)
merged_obj_plot   <- RenameIdents(merged_obj, new_labels)

umap_coords         <- data.frame(Embeddings(merged_obj_plot, "umap.rpca"))
umap_coords$cluster <- Idents(merged_obj_plot)
centroids           <- aggregate(cbind(umaprpca_1, umaprpca_2) ~ cluster, data = umap_coords, mean)

x_range <- range(umap_coords[, 1])
y_range <- range(umap_coords[, 2])

axis_origin_x <- x_range[1] - (diff(x_range) * 0.1)
axis_origin_y <- y_range[1] - (diff(y_range) * 0.1)
axis_length   <- diff(x_range) * 0.2

p <- DimPlot(merged_obj_plot,
             reduction = "umap.rpca",
             label     = FALSE,
             pt.size   = 0.1) +
  NoLegend() +
  NoAxes() +
  theme(plot.title = element_blank()) +
  geom_text_repel(data = centroids,
                  aes(x = umaprpca_1, y = umaprpca_2, label = cluster),
                  fontface          = "bold",
                  size              = 4.5,
                  color             = "black",
                  bg.color          = "white",
                  bg.r              = 0.15,
                  force             = 5,
                  min.segment.length = 0) +
  annotate("segment",
           x = axis_origin_x, xend = axis_origin_x + axis_length,
           y = axis_origin_y, yend = axis_origin_y,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
           size = 1.2, lineend = "round") +
  annotate("segment",
           x = axis_origin_x, xend = axis_origin_x,
           y = axis_origin_y, yend = axis_origin_y + axis_length,
           arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
           size = 1.2, lineend = "round") +
  annotate("text",
           x = axis_origin_x + (axis_length / 2), y = axis_origin_y - 1,
           label = "UMAP 1", hjust = 0.5, vjust = 1, fontface = "bold") +
  annotate("text",
           x = axis_origin_x - 1, y = axis_origin_y + (axis_length / 2),
           label = "UMAP 2", angle = 90, hjust = 1, vjust = 0.5, fontface = "bold") +
  annotate("text",
           x = mean(x_range), y = axis_origin_y - 2.5,
           label = paste0("Total cells = ", comma(length(Cells(merged_obj)))),
           size = 5, fontface = "bold")

print(p)

ggsave(
  filename = file.path(graphs_dir, int_method, cluster_resolution, glue("labelled_polished_{cluster_resolution}.pdf")),
  plot     = p,
  device   = cairo_pdf,
  width = 8, height = 6, dpi = 300, units = "in"
)
```

```{r}
sessionInfo()
```
