---
title: "glmmTMB"
format: html
---

Fits a negative-binomial glmmTMB model to SCT counts for every expressed gene (≥1% of cells), testing for Ploidy × Background interaction and pairwise simple effects. Runs in parallel across HPC cores. Downstream notebooks (`glmmTMB_DE.qmd`, `gsea_glmmTMB_trisomy.qmd`) read the output of this script.

> **Note:** This script is excluded from the Quarto render list — it exists as a reference and is not compiled to HTML.

**Inputs:** `output/run1/objects/prelabelled_integrated_rpca.qs2`
**Outputs:** `output/run1/objects/glmmTMB_with_interactions_corrected.qs2`

## Load libraries

```{r}
library(Seurat)
library(glmmTMB)
library(emmeans)
library(car)
library(utils)
library(qs2)
library(foreach)
library(doSNOW)
library(parallel)
library(dplyr)
```

## Load directories

```{r}
run_num <- "run1"

out_dir <- file.path("output", run_num)
graphs_dir <- file.path(out_dir, "graphs")
objects_dir <- file.path(out_dir, "objects")

dir.create(graphs_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in seurat object

```{r}
seurat_obj <- qs_read(file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))
```

## Get raw counts

Select for genes expressed in \>=1% of cells

```{r}
counts_matrix <- GetAssayData(seurat_obj, assay = "SCT", layer = "counts")

percent_expressing <- rowMeans(counts_matrix > 0)
genes_to_keep <- names(percent_expressing[percent_expressing >= 0.01])
counts_matrix <- counts_matrix[genes_to_keep, ]
```

## Set up metadata

```{r}
meta <- seurat_obj@meta.data
```

## Running glmmTMB

Core model calls a generalised linear mixed model via template model builder.

Everything in the formula is for predicting the expression for a single gene in a single cell.

Fixed effects:

Ploidy \* Background

This means that the model will calculate the main effect of ploidy, background and interaction between both of them.

Random effects are (1 \| Mouse) and (1\| Batch). This adds a random intercept for each individual object -\> tells model that cells from one group are more similar to each than the cells of the other group.

\^ Important for subtracting noise of individual mice and technical batches.

Also offset the model with a normalisation factor of the log counts.

Bigger cells have more total RNA overall, not that the genes have higher relative expression. Offset is essentially a denominator, normalising gene counts against total library size for each cell. Must be log as model's link function is logarithmic.

```{r}
# HPC Core Setup
num_cores <- 80 
cl <- makeCluster(num_cores)
registerDoSNOW(cl)

# 2. PROGRESS BAR
total_genes <- length(genes_to_keep)
pb <- txtProgressBar(max = total_genes, style = 3)

cat("Starting parallel models for", total_genes, "genes...\n")

# Create a custom progress function
progress <- function(n) {
  # \r forces R to overwrite the same line instead of spamming console
  cat(sprintf("\rProcessed %d / %d genes (%.1f%%) completed.", n, total_genes, (n/total_genes)*100))
}


opts <- list(progress = progress)

# 3. RUN THE EXTRACTION
all_results_list <- foreach(gene = genes_to_keep, 
                        .packages = c("glmmTMB", "emmeans", "car"), 
                        .export = c("counts_matrix", "meta"),
                        .options.snow = opts,
                        .errorhandling = 'pass') %dopar% {
  
  df <- data.frame(
    Counts = round(as.numeric(counts_matrix[gene, ])),
    Ploidy = meta$microglia, 
    Background = meta$app_genotype,
    Mouse = meta$orig.ident,  
    Batch = meta$pool_id,
    nCount_RNA = meta$nCount_RNA
  )
  
  # Fit Model
  model_fit <- tryCatch({
    glmmTMB(Counts ~ Ploidy * Background + (1 | Mouse) + (1 | Batch) + offset(log(nCount_RNA)),
            data = df, family = nbinom2(link = "log")) 
  }, error = function(e) return(NULL)) 
  
  if (is.null(model_fit)) return(NULL)
  
  # A. EXTRACT INTERACTION (Type III Anova)
  # This captures the significance of the "Ploidy:Background" term
  anova_res <- tryCatch({
    as.data.frame(car::Anova(model_fit, type = "III"))
  }, error = function(e) return(NULL))
  
  # B. EXTRACT SIMPLE EFFECTS (The 4 comparisons)
  emm_results <- tryCatch({
    spec1 <- pairs(emmeans(model_fit, ~ Ploidy | Background), reverse = TRUE)
    spec2 <- pairs(emmeans(model_fit, ~ Background | Ploidy), reverse = TRUE)
    rbind(as.data.frame(spec1), as.data.frame(spec2))
  }, error = function(e) return(NULL))
  
  if (is.null(emm_results) | is.null(anova_res)) return(NULL)
  
  # C. COMBINE STATS
  # Map interaction p-value from the Anova table to every contrast row for this gene
  emm_results$Gene <- gene
  emm_results$Interaction_P <- anova_res["Ploidy:Background", "Pr(>Chisq)"]
  emm_results$Ploidy_Main_P <- anova_res["Ploidy", "Pr(>Chisq)"]
  emm_results$Background_Main_P <- anova_res["Background", "Pr(>Chisq)"]
  
  # Convert Natural Log to Log2FC
  emm_results$log2FC <- emm_results$estimate / log(2)
  emm_results$FoldChange <- exp(emm_results$estimate)
  
  return(emm_results)
}

stopCluster(cl)
close(pb)

# 4. FINAL PROCESSING AND SAVE
cat("\nCombining and calculating FDR...\n")
final_df <- do.call(rbind, Filter(Negate(is.null), all_results_list))

# Calculate FDR on the pairwise p-values
final_df$FDR <- p.adjust(final_df$p.value, method = "fdr")

# Calculate FDR on the Interaction p-values (separately, as they are a different hypothesis)
# We do this per gene, not per row, to avoid over-correction
interaction_summary <- final_df %>% 
  group_by(Gene) %>% 
  summarise(Interaction_P = first(Interaction_P)) %>% 
  mutate(Interaction_FDR = p.adjust(Interaction_P, method = "fdr"))

# Join it back
final_df <- merge(final_df, interaction_summary[, c("Gene", "Interaction_FDR")], by = "Gene")

qs_save(final_df, file=file.path(objects_dir, "glmmTMB_with_interactions.qs2"))
cat("Done! Results saved.\n")
```

## Check processed output results

```{r}
final_df <- qs_read(file.path(objects_dir, "glmmTMB_with_interactions.qs2"))
```

## Changed direction of comparisons 

Trisomy - Disomy makes more sense.


```{r}
final_df_corrected <- final_df %>%
  mutate(
    # FLIP PLOIDY: If it's "B - A", flip the math
    estimate = ifelse(contrast == "B - A", estimate * -1, estimate),
    log2FC   = ifelse(contrast == "B - A", log2FC * -1, log2FC),
    FoldChange = ifelse(contrast == "B - A", 1 / FoldChange, FoldChange),
    # Then fix the label to match the new math
    contrast = ifelse(contrast == "B - A", "A - B", as.character(contrast))
)
qs_save(final_df_corrected, file.path(objects_dir, "glmmTMB_with_interactions_corrected.qs2"))

```

## Inspect corrected output

```{r}
corrected_df <- qs_read(file.path(objects_dir, "glmmTMB_with_interactions_corrected.qs2"))
corrected_df %>%
  filter(Interaction_FDR < 0.05)
```
