---
title: "glmmTMB (no chr21)"
format: html
---

Same negative-binomial glmmTMB analysis as `glmmTMB.qmd`, but excludes genes located on chromosome 21 prior to model fitting. Chromosome assignments are taken from `~/home/shared/zanettc/gene_info.csv`. Outputs are written to a `no_chr21/` subfolder so they are clearly separated from the full-genome run.

> **Note:** This script is excluded from the Quarto render list — it exists as a reference and is not compiled to HTML.

**Inputs:**
- `output/run1/objects/prelabelled_integrated_rpca.qs2`
- `~/home/shared/zanettc/gene_info.csv`

**Outputs:**
- `output/run1/objects/no_chr21/glmmTMB_with_interactions.qs2`
- `output/run1/objects/no_chr21/glmmTMB_with_interactions_corrected.qs2`

## Load libraries

```{r}
library(Seurat)
library(glmmTMB)
library(emmeans)
library(car)
library(utils)
library(qs2)
library(foreach)
library(doSNOW)
library(parallel)
library(dplyr)
library(readr)
```

## Load directories

```{r}
run_num <- "run1"

out_dir <- file.path("output", run_num)
graphs_dir <- file.path(out_dir, "graphs")
objects_dir <- file.path(out_dir, "objects")
no_chr21_dir <- file.path(objects_dir, "no_chr21")

dir.create(graphs_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(no_chr21_dir, recursive = TRUE, showWarnings = FALSE)

gene_info_path <- "~/home/shared/zanettc/gene_info.csv"
```

## Read in seurat object

```{r}
seurat_obj <- qs_read(file.path(objects_dir, "prelabelled_integrated_rpca.qs2"))
```

## Get raw counts and exclude chromosome 21 genes

Select for genes expressed in \>=1% of cells, then drop any genes annotated to chromosome 21 in `gene_info.csv`.

```{r}
counts_matrix <- GetAssayData(seurat_obj, assay = "SCT", layer = "counts")

percent_expressing <- rowMeans(counts_matrix > 0)
genes_to_keep <- names(percent_expressing[percent_expressing >= 0.01])

gene_info <- read_csv(gene_info_path, show_col_types = FALSE)
chr21_genes <- gene_info %>% filter(chrom == "21") %>% pull(gene)

n_before <- length(genes_to_keep)
genes_to_keep <- setdiff(genes_to_keep, chr21_genes)
n_after <- length(genes_to_keep)
cat(sprintf("Excluded %d chromosome 21 genes (%d -> %d genes for modelling).\n",
            n_before - n_after, n_before, n_after))

counts_matrix <- counts_matrix[genes_to_keep, ]
```

## Set up metadata

```{r}
meta <- seurat_obj@meta.data
rm(seurat_obj); gc()
```

## Running glmmTMB

Genes are processed in chunks with the worker cluster restarted between chunks. This caps cumulative per-fit memory growth (a known issue with TMB-based models) without changing the per-gene fit, so results are identical to a single-`foreach` run on the same gene set. Per-chunk results are written to `output/run1/objects/no_chr21/chunks/` so a crashed run can be resumed by re-launching — completed chunks are skipped.

```{r}
num_cores <- 80
chunk_size <- 500

chunk_dir <- file.path(no_chr21_dir, "chunks")
dir.create(chunk_dir, showWarnings = FALSE, recursive = TRUE)

total_genes <- length(genes_to_keep)
gene_chunks <- split(genes_to_keep, ceiling(seq_along(genes_to_keep) / chunk_size))
n_chunks <- length(gene_chunks)

cat(sprintf("Processing %d genes in %d chunks of up to %d (cluster restart between chunks).\n",
            total_genes, n_chunks, chunk_size))

for (i in seq_len(n_chunks)) {
  chunk_file <- file.path(chunk_dir, sprintf("chunk_%04d.qs2", i))
  if (file.exists(chunk_file)) {
    cat(sprintf("Chunk %d/%d already done — skipping.\n", i, n_chunks))
    next
  }

  cl <- makeCluster(num_cores)
  registerDoSNOW(cl)

  chunk_genes <- gene_chunks[[i]]
  t0 <- Sys.time()
  cat(sprintf("Chunk %d/%d: fitting %d genes...\n", i, n_chunks, length(chunk_genes)))

  res <- foreach(gene = chunk_genes,
                 .packages = c("glmmTMB", "emmeans", "car"),
                 .export = c("counts_matrix", "meta"),
                 .errorhandling = 'pass') %dopar% {

    df <- data.frame(
      Counts = round(as.numeric(counts_matrix[gene, ])),
      Ploidy = meta$microglia,
      Background = meta$app_genotype,
      Mouse = meta$orig.ident,
      Batch = meta$pool_id,
      nCount_RNA = meta$nCount_RNA
    )

    model_fit <- tryCatch({
      glmmTMB(Counts ~ Ploidy * Background + (1 | Mouse) + (1 | Batch) + offset(log(nCount_RNA)),
              data = df, family = nbinom2(link = "log"))
    }, error = function(e) return(NULL))

    if (is.null(model_fit)) return(NULL)

    anova_res <- tryCatch({
      as.data.frame(car::Anova(model_fit, type = "III"))
    }, error = function(e) return(NULL))

    emm_results <- tryCatch({
      spec1 <- pairs(emmeans(model_fit, ~ Ploidy | Background), reverse = TRUE)
      spec2 <- pairs(emmeans(model_fit, ~ Background | Ploidy), reverse = TRUE)
      rbind(as.data.frame(spec1), as.data.frame(spec2))
    }, error = function(e) return(NULL))

    if (is.null(emm_results) | is.null(anova_res)) return(NULL)

    emm_results$Gene <- gene
    emm_results$Interaction_P <- anova_res["Ploidy:Background", "Pr(>Chisq)"]
    emm_results$Ploidy_Main_P <- anova_res["Ploidy", "Pr(>Chisq)"]
    emm_results$Background_Main_P <- anova_res["Background", "Pr(>Chisq)"]

    emm_results$log2FC <- emm_results$estimate / log(2)
    emm_results$FoldChange <- exp(emm_results$estimate)

    return(emm_results)
  }

  stopCluster(cl)
  gc()

  qs_save(res, chunk_file)
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "mins"))
  cat(sprintf("Chunk %d/%d done in %.1f min, saved to %s\n", i, n_chunks, elapsed, chunk_file))
}

cat("\nCombining chunks and calculating FDR...\n")
all_results_list <- list()
for (i in seq_len(n_chunks)) {
  all_results_list <- c(all_results_list,
                        qs_read(file.path(chunk_dir, sprintf("chunk_%04d.qs2", i))))
}

final_df <- do.call(rbind, Filter(Negate(is.null), all_results_list))

final_df$FDR <- p.adjust(final_df$p.value, method = "fdr")

interaction_summary <- final_df %>%
  group_by(Gene) %>%
  summarise(Interaction_P = first(Interaction_P)) %>%
  mutate(Interaction_FDR = p.adjust(Interaction_P, method = "fdr"))

final_df <- merge(final_df, interaction_summary[, c("Gene", "Interaction_FDR")], by = "Gene")

qs_save(final_df, file = file.path(no_chr21_dir, "glmmTMB_with_interactions.qs2"))
cat("Done! Results saved.\n")
```

## Check processed output results

```{r}
final_df <- qs_read(file.path(no_chr21_dir, "glmmTMB_with_interactions.qs2"))
```

## Changed direction of comparisons

Trisomy - Disomy makes more sense.

```{r}
final_df_corrected <- final_df %>%
  mutate(
    estimate = ifelse(contrast == "B - A", estimate * -1, estimate),
    log2FC   = ifelse(contrast == "B - A", log2FC * -1, log2FC),
    FoldChange = ifelse(contrast == "B - A", 1 / FoldChange, FoldChange),
    contrast = ifelse(contrast == "B - A", "A - B", as.character(contrast))
)
qs_save(final_df_corrected, file.path(no_chr21_dir, "glmmTMB_with_interactions_corrected.qs2"))
```

## Inspect corrected output

```{r}
corrected_df <- qs_read(file.path(no_chr21_dir, "glmmTMB_with_interactions_corrected.qs2"))
corrected_df %>%
  filter(Interaction_FDR < 0.05)
```
